---
title: ELBO, convergence, and model selection
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ELBO, convergence, and model selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette covers the evidence lower bound (ELBO) used by mvSuSiE
for model fitting and convergence monitoring, and compares the EM and
optim algorithms for prior variance estimation.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

```{r simulate-data}
data(N3finemapping, package = "susieR")
X <- N3finemapping$X
n <- nrow(X); p <- ncol(X); r <- 10

causal <- sort(sample(p, 3))
B <- matrix(0, p, r)
for (j in causal) B[j, ] <- rnorm(r, 0, 0.5)
Y <- X %*% B + matrix(rnorm(n * r), n, r)
```

## ELBO and convergence
mvSuSiE uses the Iterative Bayesian Stepwise Selection (IBSS)
algorithm to maximize the evidence lower bound (ELBO). The ELBO
is a lower bound on the log marginal likelihood and should increase
monotonically at each iteration.

```{r fit-track-elbo}
prior <- create_mixture_prior(R = r)
fit <- mvsusie(X, Y, L = 10,
               prior_variance = prior,
               estimate_prior_variance = TRUE,
               standardize = TRUE, tol = 1e-4,
               verbosity = 0)
```

```{r elbo-plot, fig.height=2.5, fig.width=4}
elbo <- fit$elbo
cat("Iterations:", length(elbo), "\n")
cat("Final ELBO:", tail(elbo, 1), "\n")

pdat <- data.frame(iter = seq_along(elbo), elbo = elbo)
ggplot(pdat, aes(x = iter, y = elbo)) +
  geom_line(color = "royalblue") +
  geom_point(color = "royalblue", size = 2) +
  labs(x = "Iteration", y = "ELBO") +
  theme_cowplot(font_size = 12)
```

The ELBO increases monotonically, confirming correct convergence.
The `tol` parameter controls the convergence threshold: the
algorithm stops when the relative change in ELBO is less than `tol`.

## Effect pruning via log Bayes factors
mvSuSiE fits a model with $L$ single effects. Effects that do not
contribute to the model have small or negative log Bayes factors
(LBF) and are automatically pruned when `estimate_prior_variance =
TRUE`. This makes the model robust to over-specification of $L$.

```{r lbf}
cat("L =", length(fit$lbf), "single effects\n")
cat("Log Bayes factors:\n")
for (l in seq_along(fit$lbf)) {
  pruned <- if (fit$V[l] < 1e-9) " [pruned]" else ""
  cat(sprintf("  Effect %d: lbf = %8.2f, V = %.4e%s\n",
              l, fit$lbf[l], fit$V[l], pruned))
}
```

```{r lbf-plot, fig.height=2.5, fig.width=4}
lbf_dat <- data.frame(effect = seq_along(fit$lbf),
                       lbf = fit$lbf,
                       active = fit$V > 1e-9)
ggplot(lbf_dat, aes(x = effect, y = lbf, fill = active)) +
  geom_col() +
  scale_fill_manual(values = c("gray70", "royalblue"),
                    labels = c("Pruned", "Active")) +
  labs(x = "Single effect", y = "Log Bayes factor", fill = "") +
  theme_cowplot(font_size = 12)
```

## EM vs optim for prior variance estimation
mvSuSiE provides two algorithms for estimating the prior variance
scaling factor: `"EM"` (expectation-maximization) and `"optim"`
(Brent optimizer). The optim method typically converges in fewer
iterations because it directly maximizes the ELBO with respect to
the prior scaling factor at each step.

```{r em-vs-optim}
fit_em <- mvsusie(X, Y, L = 10,
                  prior_variance = prior,
                  estimate_prior_variance = TRUE,
                  estimate_prior_method = "EM",
                  standardize = TRUE, tol = 1e-4, verbosity = 0)

fit_optim <- mvsusie(X, Y, L = 10,
                     prior_variance = prior,
                     estimate_prior_variance = TRUE,
                     estimate_prior_method = "optim",
                     standardize = TRUE, tol = 1e-4, verbosity = 0)

cat("EM: ", length(fit_em$elbo), "iterations, final ELBO =",
    tail(fit_em$elbo, 1), "\n")
cat("optim:", length(fit_optim$elbo), "iterations, final ELBO =",
    tail(fit_optim$elbo, 1), "\n")
```

```{r convergence-comparison, fig.height=3, fig.width=5}
max_iter <- max(length(fit_em$elbo), length(fit_optim$elbo))
pdat <- data.frame(
  iter = c(seq_along(fit_em$elbo), seq_along(fit_optim$elbo)),
  elbo = c(fit_em$elbo, fit_optim$elbo),
  method = c(rep("EM", length(fit_em$elbo)),
             rep("optim", length(fit_optim$elbo)))
)
ggplot(pdat, aes(x = iter, y = elbo, color = method)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_color_manual(values = c("EM" = "royalblue", "optim" = "firebrick")) +
  labs(x = "Iteration", y = "ELBO", color = "Method") +
  theme_cowplot(font_size = 12)
```

The optim method reaches a similar (or higher) ELBO in fewer
iterations. Both methods should converge to the same solution,
but the optim method can be faster for models with many mixture
components where the EM update is slow to converge.

## Estimating mixture prior weights
When using a mixture prior ($K > 1$), mvSuSiE can optionally
update the mixture component weights during fitting via the
`estimate_prior_mixture_weights` parameter.

```{r mixture-weights}
fit_fixed_w <- mvsusie(X, Y, L = 10,
                       prior_variance = prior,
                       estimate_prior_variance = TRUE,
                       estimate_prior_mixture_weights = FALSE,
                       standardize = TRUE, tol = 1e-4, verbosity = 0)

fit_est_w <- mvsusie(X, Y, L = 10,
                     prior_variance = prior,
                     estimate_prior_variance = TRUE,
                     estimate_prior_mixture_weights = TRUE,
                     standardize = TRUE, tol = 1e-4, verbosity = 0)

cat("Fixed weights:", length(fit_fixed_w$elbo), "iterations, ELBO =",
    tail(fit_fixed_w$elbo, 1), "\n")
cat("Estimated weights:", length(fit_est_w$elbo), "iterations, ELBO =",
    tail(fit_est_w$elbo, 1), "\n")
```

A higher ELBO indicates a better fit to the data. Comparing the
ELBO between configurations helps decide which algorithm settings
are most appropriate for a given analysis.
