---
title: Input data formats for mvSuSiE
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Input data formats for mvSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

mvSuSiE provides three interfaces for fitting multivariate
Sum of Single Effects models, depending on the type of data available:

1. `mvsusie()` --- individual-level data (genotype matrix $X$ and
   phenotype matrix $Y$).

2. `mvsusie_ss()` --- sufficient statistics ($X^TX$, $X^TY$, $Y^TY$
   and the sample size $N$).

3. `mvsusie_rss()` --- summary statistics (*z*-scores and an LD
   matrix, or effect estimates and standard errors. Sample size $N$ is optional 
   and is fixed for all traits).

This vignette demonstrates all three interfaces on the same data set
and verifies that they produce equivalent results.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages and set the seed for reproducibility.

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

Simulate data
-------------

We use realistic genotypes from the `susieR` package and simulate a
phenotype matrix with $R = 10$ traits. Four SNPs are chosen as causal,
each with effects drawn independently across traits.

```{r simulate-data}
data(N3finemapping)
X <- N3finemapping$X
n <- nrow(X)
p <- ncol(X)
r <- 10
cat(sprintf("Genotype matrix: %d samples x %d SNPs\n", n, p))

causal <- sort(sample(p, 4))
B <- matrix(0, p, r)
for (j in causal) B[j, ] <- rnorm(r, 0, 0.5)
Y <- X %*% B + matrix(rnorm(n * r), n, r)
cat(sprintf("Phenotype matrix: %d samples x %d traits\n", n, r))
cat("Causal SNPs:", causal, "\n")
```

Set up a prior and residual variance. We use the canonical
mixture prior (which combines singleton and shared-effect covariance
matrices) and fix the residual variance to the identity (the true
value in this simulation).

```{r setup-prior}
prior <- create_mixture_prior(R = r)
resid_var <- diag(r)
```

Interface 1: Individual-level data
-----------------------------------

The most straightforward interface takes the raw genotype matrix $X$
($N \times J$) and phenotype matrix $Y$ ($N \times R$).

```{r fit-mvsusie}
fit <- mvsusie(X, Y, L = 10,
               prior_variance = prior,
               residual_variance = resid_var,
               estimate_residual_variance = FALSE,
               estimate_prior_variance = TRUE,
               standardize = TRUE,
               tol = 0.01)
```

The output is an S3 object of class `"mvsusie"` containing posterior
inclusion probabilities (PIPs), credible sets, and effect estimates.

```{r mvsusie-results}
cat("Credible sets found:", length(fit$sets$cs), "\n")
cat("PIP range:", round(range(fit$pip), 4), "\n")
fit$sets$cs
```

Interface 2: Sufficient statistics
-----------------------------------

When individual-level data are not available but cross-products can be
shared, `mvsusie_ss()` accepts the sufficient statistics $X^TX$,
$X^TY$, $Y^TY$ and the sample size $N$. The columns of $X$ and $Y$
must be centered to have mean zero before computing these
cross-products.

```{r compute-suff-stats}
X_colmeans <- colMeans(X)
Y_colmeans <- colMeans(Y)
Xc <- scale(X, center = TRUE, scale = FALSE)
Yc <- scale(Y, center = TRUE, scale = FALSE)
XtX <- crossprod(Xc)
XtY <- crossprod(Xc, Yc)
YtY <- crossprod(Yc)
```

Now fit using the sufficient statistics. Note: we also pass
`X_colmeans` and `Y_colmeans` so that the intercept can be recovered.

```{r fit-mvsusie-ss}
fit_ss <- mvsusie_ss(XtX, XtY, YtY, N = n, L = 10,
                     X_colmeans = X_colmeans,
                     Y_colmeans = Y_colmeans,
                     prior_variance = prior,
                     residual_variance = resid_var,
                     estimate_residual_variance = FALSE,
                     estimate_prior_variance = TRUE,
                     standardize = TRUE,
                     tol = 0.01)
```

Interface 3: Summary statistics (RSS)
--------------------------------------

In many genetic studies, only univariate association *z*-scores and
a linkage disequilibrium (LD) matrix are available. `mvsusie_rss()`
accepts a $J \times R$ matrix of *z*-scores and a $J \times J$
LD matrix.

```{r compute-summary-stats}
Z <- calc_z(X, Y, center = TRUE, scale = TRUE)
LD <- cor(X)
cat(sprintf("Z-score matrix: %d SNPs x %d traits\n", nrow(Z), ncol(Z)))
```

Fit the model using summary statistics. Providing the sample size `N`
is highly recommended for accurate results.

```{r fit-mvsusie-rss}
fit_rss <- mvsusie_rss(Z, LD, N = n, L = 10,
                       prior_variance = prior,
                       residual_variance = resid_var,
                       estimate_residual_variance = FALSE,
                       estimate_prior_variance = TRUE,
                       tol = 0.01)
```

Verifying agreement across interfaces
--------------------------------------

When the LD matrix and *z*-scores are computed from the same
individual-level data, all three interfaces should produce
(near-)identical results.

```{r compare-results}
cat("mvsusie vs mvsusie_ss:\n")
cat("  max |PIP diff|:", max(abs(fit$pip - fit_ss$pip)), "\n")
cat("  PIP correlation:", cor(fit$pip, fit_ss$pip), "\n\n")

cat("mvsusie vs mvsusie_rss:\n")
cat("  max |PIP diff|:", max(abs(fit$pip - fit_rss$pip)), "\n")
cat("  PIP correlation:", cor(fit$pip, fit_rss$pip), "\n")
```

The individual-level and sufficient statistics interfaces agree to
machine precision. The RSS interface shows slightly larger differences
due to the *z*-score approximation, but the results remain nearly
identical.

```{r compare-plot, fig.height=3, fig.width=7}
pdat <- data.frame(
  pip_raw = rep(fit$pip, 2),
  pip_other = c(fit_ss$pip, fit_rss$pip),
  interface = rep(c("mvsusie_ss", "mvsusie_rss"), each = p)
)
ggplot(pdat, aes(x = pip_raw, y = pip_other)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  facet_wrap(~interface) +
  labs(x = "PIP (mvsusie)", y = "PIP (other interface)") +
  theme_cowplot(font_size = 12)
```