---
title: Input data formats for mvSuSiE
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Input data formats for mvSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

mvSuSiE provides three interfaces for fitting multivariate
Sum of Single Effects models, depending on the type of data available:

1. `mvsusie()`: individual-level data (genotype matrix $X$ and
   phenotype matrix $Y$).

2. `mvsusie_ss()`: sufficient statistics ($X^TX$, $X^TY$, $Y^TY$
   and the sample size $N$).

3. `mvsusie_rss()`: summary statistics (*z*-scores and an LD
   matrix, or effect estimates and standard errors).

This vignette demonstrates all three interfaces on the same data set,
verifies equivalence, and examines the z-score-only path where
the outcome covariance `varY` is not available.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

## Simulate data

We use realistic genotypes and simulate a phenotype matrix with
$R = 10$ traits.

```{r simulate-data}
data(simdata)
X <- simdata$raw$X
n <- nrow(X)
p <- ncol(X)
r <- 10
cat(sprintf("Genotype matrix: %d samples x %d SNPs\n", n, p))

causal <- sort(sample(p, 4))
B <- matrix(0, p, r)
for (j in causal) B[j, ] <- rnorm(r, 0, 0.5)
Y <- X %*% B + matrix(rnorm(n * r), n, r)
cat(sprintf("Phenotype matrix: %d samples x %d traits\n", n, r))
cat("Causal SNPs:", causal, "\n")
```

```{r setup-prior}
prior <- create_mixture_prior(R = r)
resid_var <- diag(r)
```

## Interface 1 and 2: Individual-level and sufficient statistics

We fit both interfaces with their defaults
(`estimate_residual_variance = TRUE`).

```{r fit-individual}
fit <- mvsusie(X, Y, L = 10,
               prior_variance = prior,
               residual_variance = resid_var,
               estimate_prior_variance = TRUE,
               standardize = TRUE)
```

```{r compute-suff-stats}
X_colmeans <- colMeans(X)
Y_colmeans <- colMeans(Y)
Xc <- scale(X, center = TRUE, scale = FALSE)
Yc <- scale(Y, center = TRUE, scale = FALSE)
XtX <- crossprod(Xc)
XtY <- crossprod(Xc, Yc)
YtY <- crossprod(Yc)
```

```{r fit-ss}
fit_ss <- mvsusie_ss(XtX, XtY, YtY, N = n, L = 10,
                     X_colmeans = X_colmeans,
                     Y_colmeans = Y_colmeans,
                     prior_variance = prior,
                     residual_variance = resid_var,
                     estimate_prior_variance = TRUE,
                     standardize = TRUE)
```

```{r compare-ind-ss}
cat("mvsusie vs mvsusie_ss (both defaults):\n")
cat("  max |PIP diff|:", max(abs(fit$pip - fit_ss$pip)), "\n")
cat("  max |coef diff|:", max(abs(coef(fit)[-1, ] - coef(fit_ss)[-1, ])), "\n")
```

The PIP and coefficient differences are at machine precision,
confirming that `mvsusie()` and `mvsusie_ss()` are numerically
identical when given the same data.

## Interface 3a: RSS with full information

When effect estimates (`Bhat`), standard errors (`Shat`), and the
sample covariance of Y (`varY`) are all provided alongside the LD
matrix and sample size, `mvsusie_rss()` reconstructs the exact
sufficient statistics and produces results identical to `mvsusie()`
and `mvsusie_ss()`.

```{r compute-bhat-shat}
Bhat <- matrix(0, p, r)
Shat <- matrix(0, p, r)
for (outcome in seq_len(r)) {
  reg <- susieR:::univariate_regression(Xc, Yc[, outcome])
  Bhat[, outcome] <- reg$betahat
  Shat[, outcome] <- reg$sebetahat
}
LD <- cor(X)
varY_mat <- cov(Y)
```

```{r fit-rss-bhat}
fit_rss_bhat <- mvsusie_rss(Bhat = Bhat, Shat = Shat, R = LD,
                             N = n, varY = varY_mat, L = 10,
                             prior_variance = prior,
                             residual_variance = resid_var,
                             estimate_residual_variance = TRUE,
                             estimate_prior_variance = TRUE)
```

```{r compare-rss-bhat}
cat("mvsusie vs mvsusie_rss (Bhat/Shat/varY):\n")
cat("  max |PIP diff|:", max(abs(fit$pip - fit_rss_bhat$pip)), "\n")
cat("  max |coef diff|:", max(abs(coef(fit)[-1, ] - coef(fit_rss_bhat)[-1, ])), "\n")
```

With `Bhat`, `Shat`, and `varY` all provided, `mvsusie_rss()`
reconstructs the exact sufficient statistics internally, so the PIPs
and coefficients match `mvsusie()` to machine precision.

Similarly, when z-scores are provided together with `varY`, the
PIPs also match closely. (The coefficients are on a standardized
scale in this case, so only PIPs are compared.)

```{r fit-rss-z-vary}
Z <- calc_z(X, Y, center = TRUE, scale = TRUE)
fit_rss_zv <- mvsusie_rss(Z, LD, N = n, varY = varY_mat, L = 10,
                           prior_variance = prior,
                           residual_variance = resid_var,
                           estimate_residual_variance = TRUE,
                           estimate_prior_variance = TRUE)
```

```{r compare-rss-z-vary}
cat("mvsusie vs mvsusie_rss (Z + varY):\n")
cat("  max |PIP diff|:", max(abs(fit$pip - fit_rss_zv$pip)), "\n")
```

The z-score path with `varY` also produces PIPs that match at near
machine precision.  Note that the coefficients are on a standardized
scale in this case (they differ from the individual-level coefficients
by a factor related to $\sqrt{N}$), so only PIP agreement is checked.

```{r compare-3a-plot, fig.height=3, fig.width=7}
pdat <- data.frame(
  pip_ind = rep(fit$pip, 2),
  pip_rss = c(fit_rss_bhat$pip, fit_rss_zv$pip),
  interface = rep(c("Bhat/Shat/varY", "Z + varY"), each = p)
)
ggplot(pdat, aes(x = pip_ind, y = pip_rss)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  facet_wrap(~interface) +
  labs(x = "PIP (mvsusie)", y = "PIP (mvsusie_rss)") +
  theme_cowplot(font_size = 12)
```

## Interface 3b: RSS with z-scores only (no varY)

In the most common summary-data scenario, only z-scores, the LD
matrix, and sample size are available --- without `varY` or
`Bhat`/`Shat`. In this setting, the outcome correlation `cor(Y)`
needed for the $Y^TY$ matrix is estimated internally from near-null
z-scores (SNPs with $\max|Z| < 2$). This estimate captures the
residual correlation structure but may miss signal-induced
cross-trait correlations.

The default for this path is `estimate_residual_variance = FALSE`,
following the recommendation in
[Zou *et al.* (2022)][susie-rss] that the residual variance is
difficult to estimate reliably from summary statistics alone.

```{r fit-rss-z-false}
fit_rss_f <- mvsusie_rss(Z, LD, N = n, L = 10,
                          prior_variance = prior,
                          residual_variance = resid_var,
                          estimate_residual_variance = FALSE,
                          estimate_prior_variance = TRUE)
```

```{r fit-rss-z-true}
fit_rss_t <- mvsusie_rss(Z, LD, N = n, L = 10,
                          prior_variance = prior,
                          residual_variance = resid_var,
                          estimate_residual_variance = TRUE,
                          estimate_prior_variance = TRUE)
```

```{r compare-3b}
cat("=== Z-score only: FALSE vs TRUE ===\n")
cat("  Credible sets (FALSE):", length(fit_rss_f$sets$cs), "\n")
cat("  Credible sets (TRUE):", length(fit_rss_t$sets$cs), "\n")
cat("  PIP correlation:", round(cor(fit_rss_f$pip, fit_rss_t$pip), 4), "\n\n")

cat("=== Z-score only vs individual-level (defaults) ===\n")
cat("mvsusie (TRUE) vs mvsusie_rss z-only (FALSE):\n")
cat("  max |PIP diff|:", round(max(abs(fit$pip - fit_rss_f$pip)), 4), "\n")
cat("  PIP correlation:", round(cor(fit$pip, fit_rss_f$pip), 4), "\n\n")

cat("mvsusie (TRUE) vs mvsusie_rss z-only (TRUE):\n")
cat("  max |PIP diff|:", round(max(abs(fit$pip - fit_rss_t$pip)), 4), "\n")
cat("  PIP correlation:", round(cor(fit$pip, fit_rss_t$pip), 4), "\n\n")

cat("=== Z-score only vs Z + varY (both TRUE) ===\n")
cat("  max |PIP diff|:", round(max(abs(fit_rss_zv$pip - fit_rss_t$pip)), 4), "\n")
cat("  PIP correlation:", round(cor(fit_rss_zv$pip, fit_rss_t$pip), 4), "\n")
```

With the default `estimate_residual_variance = FALSE`, the z-score-only
path gives PIPs that are nearly identical to the individual-level fit
(PIP correlation $\approx 1$).  Enabling residual variance estimation
(`TRUE`) without `varY` can lead to noticeably different results: the
estimated outcome correlation from null z-scores may not perfectly
capture the true `cor(Y)`, causing the iterative residual variance
updates to drift.  In this simulation the `TRUE` path finds an
additional credible set and the PIP correlation with the
individual-level fit drops substantially.

This illustrates the recommendation from
[Zou *et al.* (2022)][susie-rss]: when only z-scores and LD are
available, fixing the residual variance
(`estimate_residual_variance = FALSE`) is the safer default.

```{r compare-3b-plot, fig.height=3, fig.width=10}
pdat <- data.frame(
  pip_ref = rep(fit$pip, 3),
  pip_rss = c(fit_rss_f$pip, fit_rss_t$pip, fit_rss_zv$pip),
  label = rep(c("Z only (FALSE)", "Z only (TRUE)", "Z + varY (TRUE)"),
              each = p)
)
ggplot(pdat, aes(x = pip_ref, y = pip_rss)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  facet_wrap(~label) +
  labs(x = "PIP (mvsusie, default)", y = "PIP (mvsusie_rss)") +
  theme_cowplot(font_size = 12)
```

The left panel (`FALSE`, the default) shows tight agreement with the
individual-level fit.  The middle panel (`TRUE`, no `varY`) shows
scatter away from the diagonal --- the residual variance estimator
is unreliable without `varY`.  The right panel (`TRUE` with `varY`)
recovers the individual-level PIPs, confirming that the divergence in
the middle panel is due to the missing `varY`, not the estimation
procedure itself.

## Coefficient estimates across interfaces

The z-score path works on a standardized scale: internally,
$X^TX = (N-1)\,R$ and $X^TY = \sqrt{N-1}\,Z$, so the posterior mean
coefficients differ from the individual-level coefficients by a
factor related to $\sqrt{N}$.  Despite the scale difference, when
both interfaces use the same `estimate_residual_variance` setting,
the coefficients should be perfectly correlated because the
underlying model is the same up to a constant rescaling.

To isolate this, we re-fit `mvsusie()` with
`estimate_residual_variance = FALSE` so that it matches the RSS
default:

```{r fit-ind-false}
fit_f <- mvsusie(X, Y, L = 10,
                 prior_variance = prior,
                 residual_variance = resid_var,
                 estimate_residual_variance = FALSE,
                 estimate_prior_variance = TRUE,
                 standardize = TRUE)
```

```{r coef-compare}
coef_ind_true  <- coef(fit)[-1, ]       # individual, TRUE (default)
coef_ind_false <- coef(fit_f)[-1, ]     # individual, FALSE
coef_rss_false <- coef(fit_rss_f)[-1, ] # RSS z-only, FALSE (default)

cat("=== Coefficient correlations (vectorized over all J x R entries) ===\n\n")
cat("Both FALSE (ind vs RSS z-only):\n")
cat("  correlation:", round(cor(as.vector(coef_ind_false),
                                as.vector(coef_rss_false)), 6), "\n\n")
cat("ind TRUE vs RSS z-only FALSE (both defaults):\n")
cat("  correlation:", round(cor(as.vector(coef_ind_true),
                                as.vector(coef_rss_false)), 6), "\n")
```

When both interfaces use the same `estimate_residual_variance`
setting (`FALSE`), the coefficients are strongly correlated even though
they live on different scales.  The z-score path coefficients are
roughly $1/\sqrt{N}$ times the individual-level coefficients, because
the sufficient statistics are constructed from standardized quantities
($X^TX = (N-1)\,R$, $X^TY \propto Z$).  Comparing the defaults ---
individual-level with `TRUE` and RSS with `FALSE` --- gives a similar
correlation, because the residual variance estimation primarily
affects convergence and the ELBO, with only a modest effect on the
posterior means.

```{r coef-plot, fig.height=3, fig.width=7}
pdat <- data.frame(
  coef_ind = c(as.vector(coef_ind_false), as.vector(coef_ind_true)),
  coef_rss = rep(as.vector(coef_rss_false), 2),
  setting  = rep(c("Both FALSE", "ind TRUE vs RSS FALSE (defaults)"),
                 each = length(coef_ind_false))
)
ggplot(pdat, aes(x = coef_ind, y = coef_rss)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  facet_wrap(~setting, scales = "free") +
  labs(x = "Coefficient (mvsusie)", y = "Coefficient (mvsusie_rss, z-only)") +
  theme_cowplot(font_size = 12)
```

In the left panel (both `FALSE`), the coefficients fall along a line
with slope less than one, reflecting the standardized scale of the
z-score path.  In the right panel (default settings: individual `TRUE`
vs RSS `FALSE`), the pattern is similar --- the individual-level
coefficients are slightly larger because re-estimating the residual
variance adjusts the posterior shrinkage, but the overall agreement
remains strong.

For applications where the coefficient magnitude matters (e.g.,
polygenic prediction), `mvsusie_rss()` with `Bhat`/`Shat`/`varY`
(Interface 3a) is preferred, as it produces coefficients on the
original scale identical to `mvsusie()`.

[susie-rss]: https://doi.org/10.1371/journal.pgen.1010299
