---
title: Scaling mvSuSiE to many traits
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scaling mvSuSiE to many traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette discusses strategies for running mvSuSiE with a large
number of traits ($R = 100$). With many traits, computation time
and memory usage become important considerations.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(ggplot2)
library(cowplot)
```

## Key considerations for large R
When the number of traits $R$ is large (e.g., $R > 50$):

1. **Prior specification**: A canonical or data-driven mixture prior
   is recommended. For very large $R$, the number of canonical prior
   matrices grows, so data-driven priors from mashr that select only
   the relevant components are preferred.

2. **Precompute cache**: Setting `precompute_cache = TRUE` (default)
   caches eigendecompositions of the prior covariance matrices,
   providing substantial speedups for large $R$. The trade-off is
   increased memory usage.

3. **Residual variance estimation**: When $N < R$, the sample
   covariance of $Y$ is singular. In this case, regularization
   methods such as the `flashier` package can be used to estimate
   a low-rank plus diagonal residual covariance.

4. **C++ acceleration**: The core computations are implemented in C++
   for efficiency. Multi-threading is available via the `n_thread`
   argument.

## Example: R = 100
The following code shows how to set up and run an analysis with 100
traits. Due to the computational cost, we show the code but do not
execute it during the vignette build.

```{r large-r-setup, eval=FALSE}
library(susieR)

# Use realistic genotypes
data(N3finemapping)
X <- N3finemapping$X  # 574 x 1001
n <- nrow(X); p <- ncol(X); r <- 100

# Simulate 100 traits with shared causal variants
set.seed(42)
B <- matrix(0, p, r)
causal <- sort(sample(p, 4))
for (j in causal) B[j, ] <- rnorm(r, 0, 0.3)
Y <- X %*% B + matrix(rnorm(n * r), n, r)

# Canonical prior (105 components for R=100)
prior <- create_mixture_prior(R = r)

# Fit with precompute_cache and timing
t0 <- proc.time()
fit <- mvsusie(X, Y, L = 10,
               prior_variance = prior,
               estimate_prior_variance = TRUE,
               precompute_cache = TRUE,
               standardize = TRUE,
               tol = 0.01, n_thread = 4)
timing <- (proc.time() - t0)[["elapsed"]]
```

**Benchmark results.** On an AMD Ryzen 5 7600X (6-core, 12 threads)
with 64 GB RAM, the above analysis with $N = 574$ samples, $J = 1001$
SNPs, $R = 100$ traits, and 105 canonical prior components completed
in approximately **62 seconds** (5 IBSS iterations) with
`precompute_cache = TRUE`. The fitted model object uses approximately
41 MB of memory.

Without `precompute_cache`, the same analysis takes approximately
67 seconds---a modest difference in this case because the canonical
prior matrices share structure that limits the cache benefit. The
speedup from `precompute_cache` is more pronounced with data-driven
priors that have diverse covariance structures.
