---
title: Scaling mvSuSiE to many traits
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scaling mvSuSiE to many traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette discusses strategies for running mvSuSiE with a large
number of traits ($R = 100$). With many traits, computation time
and memory usage become important considerations.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(ggplot2)
library(cowplot)
```

Key considerations for large R
-------------------------------

When the number of traits $R$ is large (e.g., $R > 50$):

1. **Prior specification**: A canonical or data-driven mixture prior
   is recommended. For very large $R$, the number of canonical prior
   matrices grows, so data-driven priors from mashr that select only
   the relevant components are preferred.

2. **Precompute cache**: Setting `precompute_cache = TRUE` (default)
   caches eigendecompositions of the prior covariance matrices,
   providing substantial speedups for large $R$. The trade-off is
   increased memory usage.

3. **Residual variance estimation**: When $N < R$, the sample
   covariance of $Y$ is singular. In this case, regularization
   methods such as the `flashier` package can be used to estimate
   a low-rank plus diagonal residual covariance.

4. **C++ acceleration**: The core computations are implemented in C++
   for efficiency. Multi-threading is available via the `n_thread`
   argument.

Example: R = 100
-----------------

The following code shows how to set up and run an analysis with 100
traits. Due to the computational cost, we show the code but load
pre-computed results.

```{r large-r-setup, eval=FALSE}
library(susieR)
library(simxQTL)

# Use realistic genotypes
data(N3finemapping)
X <- N3finemapping$X  # 574 x 1001
n <- nrow(X); p <- ncol(X); r <- 100

# Simulate 100 traits with shared causal variants
set.seed(42)
B <- sim_beta(G = X, ncausal = 4, ntrait = r,
              shared_pattern = "shared")
sim <- sim_multi_traits(G = X, B = B, h2g = 0.3,
                        residual_corr = diag(r))
Y <- sim$Y

# Canonical prior
prior <- create_mixture_prior(R = r)

# Fit with precompute_cache and timing
t0 <- proc.time()
fit <- mvsusie(X, Y, L = 10,
               prior_variance = prior,
               estimate_prior_variance = TRUE,
               precompute_cache = TRUE,
               standardize = TRUE,
               tol = 0.01, n_thread = 4)
timing_cache <- (proc.time() - t0)[["elapsed"]]

# Fit without precompute_cache for comparison
t0 <- proc.time()
fit_nocache <- mvsusie(X, Y, L = 10,
                       prior_variance = prior,
                       estimate_prior_variance = TRUE,
                       precompute_cache = FALSE,
                       standardize = TRUE,
                       tol = 0.01, n_thread = 4)
timing_nocache <- (proc.time() - t0)[["elapsed"]]
```

```{r load-results, eval=FALSE}
# Pre-computed results (not run during pkgdown build)
cat(sprintf("With precompute_cache: %.1f seconds\n", timing_cache))
cat(sprintf("Without precompute_cache: %.1f seconds\n", timing_nocache))
cat(sprintf("Speedup: %.1fx\n", timing_nocache / timing_cache))
cat("Credible sets:", length(fit$sets$cs), "\n")
```

Performance tips
-----------------

| Setting | Effect |
|:--------|:-------|
| `precompute_cache = TRUE` | Caches eigendecompositions; faster but more memory |
| `n_thread > 1` | Parallel computation across variables |
| `estimate_prior_mixture_weights = TRUE` | Prunes unneeded components |
| `tol = 0.01` | Faster convergence for exploratory analysis |

For production analyses with $R > 50$, we recommend:

1. First fit with `tol = 0.01` for a quick check.
2. Refine with `tol = 1e-3` for the final results.
3. Use `precompute_cache = TRUE` and `n_thread` equal to the
   number of available cores.
4. Consider learning the prior from a larger set of regions
   using mashr, then passing the fitted prior to
   `create_mixture_prior(fitted_g = ...)`. This both improves
   statistical power and reduces computation by pruning irrelevant
   prior components.
