---
title: Scaling mvSuSiE to many traits
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scaling mvSuSiE to many traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates running mvSuSiE with a relatively large number of
traits ($R = 100$) and provides benchmark timings for typical
configurations.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(ggplot2)
library(cowplot)
```

## Key considerations for large R

When the number of traits $R$ is large (e.g., $R > 50$):

1. **Prior specification**: A canonical or data-driven mixture prior
   is recommended. For very large $R$, the number of canonical prior
   matrices grows, so data-driven priors from mashr that select only
   the relevant components are preferred.

2. **Precompute cache**: Setting `precompute_cache = TRUE` (default)
   caches eigendecompositions of the prior covariance matrices,
   providing substantial speedups for large $R$. The trade-off is
   increased memory usage.

3. **Residual variance estimation**: When $N < R$, the sample
   covariance of $Y$ is singular. In this case, regularization methods 
   [such as using `flashier` package](https://github.com/stephenslab/mr.mashr/blob/be538c3856d12a127104058708ba58e0af2da670/R/misc.R#L272) 
   can be used to estimate a low-rank plus diagonal residual covariance,
   and specified into mvSuSiE via `residual_variance = ...`. If still left to `NULL`,
   `mvsusie` uses pairwise complete covariance estimation with a positive
   definiteness correction as the initial residual covariance.

4. **C++ acceleration**: The core computations are implemented in Rcpp
   for efficiency and is available by default. Multi-threading is available via the `n_thread`
   argument.

## Example: R = 100

```{r large-r-setup}
data(simdata)
X <- simdata$raw$X
n <- nrow(X); p <- ncol(X); r <- 100

# Simulate 100 traits with shared causal variants
set.seed(42)
B <- matrix(0, p, r)
causal <- sort(sample(p, 4))
for (j in causal) B[j, ] <- rnorm(r, 0, 0.3)
Y <- X %*% B + matrix(rnorm(n * r), n, r)
cat(sprintf("Data: %d samples, %d SNPs, %d traits\n", n, p, r))
cat("Causal SNPs:", causal, "\n")

# Canonical prior (105 components for R=100)
prior <- create_mixture_prior(R = r)
cat("Prior components:", prior$n_component, "\n")
```

```{r large-r-fit}
t0 <- proc.time()
fit <- mvsusie(X, Y, L = 10,
               prior_variance = prior)
timing <- (proc.time() - t0)[["elapsed"]]
cat(sprintf("Elapsed time: %.1f seconds\n", timing))
cat(sprintf("IBSS iterations: %d\n", length(fit$elbo)))
cat(sprintf("Credible sets: %d\n", length(fit$sets$cs)))
```

```{r large-r-results}
# Check which causal SNPs are captured
for (cs_name in names(fit$sets$cs)) {
  cs_vars <- fit$sets$cs[[cs_name]]
  captured <- intersect(cs_vars, causal)
  cat(sprintf("%s: %d SNPs, captures causal %s\n",
              cs_name, length(cs_vars),
              if (length(captured) > 0) paste(captured, collapse = ", ")
              else "none"))
}
```

```{r large-r-pip, fig.height=2, fig.width=6}
pdat <- data.frame(pos = seq_len(p), pip = fit$pip)
ggplot(pdat, aes(x = pos, y = pip)) +
  geom_point(shape = 20, size = 1, color = "royalblue", alpha = 0.5) +
  geom_vline(xintercept = causal, linetype = "dashed",
             color = "red", alpha = 0.5) +
  labs(x = "SNP position", y = "PIP") +
  theme_cowplot(font_size = 12)
```