---
title: Distinguishing linkage from pleiotropy
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distinguishing linkage from pleiotropy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A key advantage of multi-trait fine-mapping is the ability to
distinguish between *pleiotropy* (one SNP affecting multiple traits)
and *linkage* (different correlated SNPs each affecting different
traits). When two correlated SNPs have distinct trait-specific
effects, mvSuSiE can separate them into different credible sets.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

Simulation setup
-----------------

We use realistic genotypes and select two correlated SNPs
(approximately 70% correlation) as causal variants. Each causal
SNP affects a different subset of traits.

```{r simulate-data}
data(N3finemapping)
X <- N3finemapping$X
n <- nrow(X); p <- ncol(X); r <- 10

# Find a pair of SNPs with ~70% correlation
R_mat <- cor(X)
pairs <- which(abs(R_mat) > 0.65 & abs(R_mat) < 0.75 &
               upper.tri(R_mat), arr.ind = TRUE)
snp1 <- pairs[1, 1]
snp2 <- pairs[1, 2]
cat(sprintf("Causal SNPs: %d and %d (correlation: %.3f)\n",
            snp1, snp2, R_mat[snp1, snp2]))
```

SNP 1 affects traits 1--5, SNP 2 affects traits 6--10. This
creates a linkage scenario: two correlated variants with
non-overlapping effects.

```{r create-effects}
B <- matrix(0, p, r)
B[snp1, 1:5] <- rnorm(5, 0, 1)    # SNP 1 -> traits 1-5
B[snp2, 6:10] <- rnorm(5, 0, 1)   # SNP 2 -> traits 6-10
Y <- X %*% B + matrix(rnorm(n * r), n, r)
```

Fitting mvSuSiE with a mixture prior
--------------------------------------

A mixture prior with block-diagonal structure is well-suited for
detecting distinct effect patterns. The canonical prior generated
by `create_mixture_prior()` includes both shared and singleton
patterns.

```{r fit-model}
prior <- create_mixture_prior(R = r)
fit <- mvsusie(X, Y, L = 10,
               prior_variance = prior,
               estimate_prior_variance = TRUE,
               standardize = TRUE, tol = 0.01)
```

Credible sets and separation
-----------------------------

```{r credible-sets}
cat("Number of credible sets:", length(fit$sets$cs), "\n\n")
for (cs_name in names(fit$sets$cs)) {
  cs_vars <- fit$sets$cs[[cs_name]]
  has_snp1 <- snp1 %in% cs_vars
  has_snp2 <- snp2 %in% cs_vars
  cat(sprintf("%s: %d SNPs, contains SNP%d=%s SNP%d=%s\n",
              cs_name, length(cs_vars),
              snp1, has_snp1, snp2, has_snp2))
}
```

If the two causal SNPs appear in *different* credible sets, mvSuSiE
has successfully distinguished linkage from pleiotropy.

Trait-specific effects
-----------------------

```{r effect-plot, fig.height=3.5, fig.width=3.5}
out <- mvsusie_plot(fit, add_cs = TRUE, lfsr_cutoff = 0.05)
out$effect_plot
```

The effect plot should show that the credible sets have distinct
trait-specific patterns, confirming the separation of the two
linked signals.

Comparison with univariate analysis
-------------------------------------

A univariate (single-trait) analysis cannot distinguish which
SNP is causal for which trait when the SNPs are correlated.

```{r univariate-comparison}
# Fit per-trait susie and check if causal SNPs are in CS
for (t in c(1, 6)) {
  fit_uv <- susie(X, Y[, t], L = 10, standardize = TRUE)
  if (length(fit_uv$sets$cs) > 0) {
    cs1 <- fit_uv$sets$cs[[1]]
    cat(sprintf("Trait %d CS: %d SNPs, contains SNP%d=%s, SNP%d=%s\n",
                t, length(cs1),
                snp1, snp1 %in% cs1,
                snp2, snp2 %in% cs1))
  } else {
    cat(sprintf("Trait %d: no CS found\n", t))
  }
}
```

In the univariate analysis, the credible sets for different traits
may each contain *both* correlated SNPs, providing no resolution.
The multivariate analysis can separate them by leveraging the
distinct trait patterns.
