---
title: Distinguishing linkage from pleiotropy
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distinguishing linkage from pleiotropy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A key advantage of multi-trait fine-mapping is the ability to
distinguish between *pleiotropy* (one SNP affecting multiple traits)
and *linkage* (different correlated SNPs each affecting different
traits). When two correlated SNPs have distinct trait-specific
effects, mvSuSiE can separate them into different credible sets,
resolving *LD-induced pleiotropy*.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

## LD-induced pleiotropy: the problem
Consider two causal SNPs that are in *negative* LD. SNP 1 increases
traits 1--5, and SNP 2 increases traits 6--10. Because of the
negative LD, the marginal z-scores show associations in *all 10
traits* for *both* SNPs---positive for the traits they truly
affect and negative for the other traits (due to the LD partner's
effect). This creates the appearance of broad pleiotropy when in
reality each SNP is trait-specific.

```{r simulate-data}
data(N3finemapping)
X <- N3finemapping$X
n <- nrow(X); p <- ncol(X); r <- 10

# Find a pair of SNPs with strong negative LD (~-0.7)
R_mat <- cor(X)
pairs <- which(R_mat < -0.6 & R_mat > -0.8 &
               upper.tri(R_mat), arr.ind = TRUE)
snp1 <- pairs[1, 1]
snp2 <- pairs[1, 2]
cat(sprintf("Causal SNPs: %d and %d (correlation: %.3f)\n",
            snp1, snp2, R_mat[snp1, snp2]))
```

Each causal SNP has *positive* effects on its own 5 traits:

```{r create-effects}
B <- matrix(0, p, r)
effect_size <- 1.5
B[snp1, 1:5] <- abs(rnorm(5, effect_size, 0.2))   # SNP 1 -> traits 1-5
B[snp2, 6:10] <- abs(rnorm(5, effect_size, 0.2))   # SNP 2 -> traits 6-10
Y <- X %*% B + matrix(rnorm(n * r), n, r)

cat("True effects of SNP", snp1, "(traits 1-5):",
    round(B[snp1, 1:5], 2), "\n")
cat("True effects of SNP", snp2, "(traits 6-10):",
    round(B[snp2, 6:10], 2), "\n")
```

The marginal z-scores show associations in all 10 traits for both
SNPs, due to the negative LD confounding:

```{r marginal-z-scores, fig.height=3, fig.width=6}
Z <- calc_z(X, Y, center = TRUE, scale = TRUE)
z_dat <- data.frame(
  trait = rep(paste0("T", 1:r), 2),
  zscore = c(Z[snp1, ], Z[snp2, ]),
  snp = rep(c(paste("SNP", snp1), paste("SNP", snp2)), each = r)
)
z_dat$trait <- factor(z_dat$trait, levels = paste0("T", 1:r))

ggplot(z_dat, aes(x = trait, y = zscore, fill = snp)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values = c("royalblue", "firebrick")) +
  labs(x = "Trait", y = "Z-score", fill = "") +
  theme_cowplot(font_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Both SNPs show z-scores in all 10 traits. Without multi-trait
fine-mapping, one might conclude that both SNPs are broadly
pleiotropic. But the *sign pattern*---positive in 5 traits and
negative in the other 5---is a hallmark of LD-induced pleiotropy.

## Resolving with mvSuSiE
mvSuSiE with a mixture prior can separate the two linked signals
into different credible sets, each with its own trait-specific
effect pattern.

```{r fit-model}
prior <- create_mixture_prior(R = r)
fit <- mvsusie(X, Y, L = 10,
               prior_variance = prior,
               estimate_prior_variance = TRUE,
               standardize = TRUE, tol = 0.01)
```

```{r credible-sets}
cat("Number of credible sets:", length(fit$sets$cs), "\n\n")
for (cs_name in names(fit$sets$cs)) {
  cs_vars <- fit$sets$cs[[cs_name]]
  has_snp1 <- snp1 %in% cs_vars
  has_snp2 <- snp2 %in% cs_vars
  cat(sprintf("%s: %d SNPs, contains SNP%d=%s SNP%d=%s\n",
              cs_name, length(cs_vars),
              snp1, has_snp1, snp2, has_snp2))
}
```

```{r effect-plot, fig.height=3.5, fig.width=3.5}
out <- mvsusie_plot(fit, add_cs = TRUE, lfsr_cutoff = 0.05)
out$effect_plot
```

The effect plot shows that each credible set has effects in only
5 traits---exactly matching the true simulation. mvSuSiE has
correctly resolved the LD-induced pleiotropy into two
trait-specific signals.

```{r pip-plot, fig.height=2, fig.width=6}
out$pip_plot
```

## Comparison with univariate analysis
A univariate (single-trait) analysis cannot distinguish which
SNP is causal for which trait when the SNPs are correlated.

```{r univariate-comparison}
for (t in c(1, 6)) {
  fit_uv <- susie(X, Y[, t], L = 10, standardize = TRUE)
  if (length(fit_uv$sets$cs) > 0) {
    cs1 <- fit_uv$sets$cs[[1]]
    cat(sprintf("Trait %d CS: %d SNPs, contains SNP%d=%s, SNP%d=%s\n",
                t, length(cs1),
                snp1, snp1 %in% cs1,
                snp2, snp2 %in% cs1))
  } else {
    cat(sprintf("Trait %d: no CS found\n", t))
  }
}
```

In the univariate analysis, the credible sets for different traits
may each contain *both* correlated SNPs, providing no resolution.
The multivariate analysis leverages the distinct trait patterns to
separate the two linked signals.
