---
title: Distinguishing linkage from pleiotropy
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distinguishing linkage from pleiotropy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In multi-trait fine-mapping it is possible in principle, though not always in practice,
to distinguish between *pleiotropy* (one SNP affecting multiple traits)
and *linkage* (different correlated SNPs each affecting different
traits). When two correlated SNPs have distinct trait-specific
effects, if the analysis is sufficiently powered, 
mvSuSiE can separate them into different credible sets,
resolving *LD-induced pleiotropy*.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(2)
```

## LD-induced pleiotropy: the problem

Consider two causal SNPs in *negative* LD ($r \approx -0.66$).
SNP 1 affects traits 1--5 and SNP 2 affects traits 6--10. Because
of the negative LD, the marginal association statistics show signal
in *all 10 traits* for *both* SNPs, positive for the traits they
truly affect and negative for the other traits (due to their "LD buddies"). 
This creates the appearance of broad pleiotropy
when in reality each SNP is trait-specific.

```{r simulate-data}
data(simdata)
X <- simdata$raw$X
n <- nrow(X); p <- ncol(X); r <- 10

# Find a pair of SNPs with strong negative LD
R_mat <- cor(X)
pairs <- which(R_mat < -0.6 & R_mat > -0.8 &
               upper.tri(R_mat), arr.ind = TRUE)
snp1 <- pairs[1, 1]
snp2 <- pairs[1, 2]
cat(sprintf("Causal SNPs: %d and %d (LD r = %.3f)\n",
            snp1, snp2, R_mat[snp1, snp2]))
```

We set per-trait PVE (heritability) at $\approx 7\text{--}9\%$,
consistent with a moderately strong eQTL signal,

```{r create-effects}
B <- matrix(0, p, r)
effect_size <- 2.0
B[snp1, 1:5] <- effect_size    # SNP 1 -> traits 1-5
B[snp2, 6:10] <- effect_size   # SNP 2 -> traits 6-10
Y <- X %*% B + matrix(rnorm(n * r), n, r)

# Per-trait PVE
pve <- sapply(1:r, function(t) {
  var(X %*% B[, t]) / var(Y[, t])
})
cat(sprintf("Per-trait PVE range: %.1f%% -- %.1f%%\n",
            100 * min(pve), 100 * max(pve)))
```

### Marginal associations are misleading

The marginal z-scores and p-values show associations across almost all
10 traits for both SNPs, due to the negative LD confounding:

```{r marginal-z-scores, fig.height=3.5, fig.width=7}
Z <- calc_z(X, Y, center = TRUE, scale = TRUE)
pval <- 2 * pnorm(-abs(Z))

z_dat <- data.frame(
  trait = rep(paste0("T", 1:r), 2),
  zscore = c(Z[snp1, ], Z[snp2, ]),
  pval = c(pval[snp1, ], pval[snp2, ]),
  snp = rep(c(paste("SNP", snp1), paste("SNP", snp2)), each = r)
)
z_dat$trait <- factor(z_dat$trait, levels = paste0("T", 1:r))
z_dat$direction <- factor(ifelse(z_dat$zscore > 0, "+1", "-1"),
                          levels = c("-1", "+1"))
z_dat$logp <- -log10(z_dat$pval)

ggplot(z_dat, aes(x = snp, y = trait, fill = direction, size = logp)) +
  geom_point(shape = 21, stroke = 0.5, color = "white") +
  scale_fill_manual(values = c("darkblue", "red"), drop = FALSE) +
  scale_size(range = c(1, 6),
             breaks = unname(round(quantile(z_dat$logp,
                                            seq(0, 1, length.out = 4)), 1))) +
  labs(x = "", y = "", fill = "direction", size = expression(-log[10](p))) +
  guides(fill = guide_legend(override.aes = list(size = 2)),
         size = guide_legend(override.aes = list(shape = 21, fill = "black",
                                                  color = "white",
                                                  stroke = 0.5))) +
  theme_cowplot(font_size = 10) +
  background_grid(major = "xy", minor = "none",
                  size.major = 0.3, colour.major = "grey85") +
  panel_border(colour = "grey70") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Both SNPs show significant associations in most of the 10
traits. All of them have $p<0.05$. Without multi-trait fine-mapping, one might conclude
that both SNPs are broadly pleiotropic, with positive effects in 5 traits and negative in the other 5.

## Resolving with mvSuSiE

mvSuSiE with a mixture prior can separate the two linked signals
into different credible sets, each with its own trait-specific
effect pattern.

```{r fit-model}
prior <- create_mixture_prior(R = r)
fit <- mvsusie(X, Y, L = 10,
               prior_variance = prior)
```

```{r credible-sets}
cat("Number of credible sets:", length(fit$sets$cs), "\n\n")
for (cs_name in names(fit$sets$cs)) {
  cs_vars <- fit$sets$cs[[cs_name]]
  l_idx <- as.integer(gsub("L", "", cs_name))
  sig_traits <- which(fit$single_effect_lfsr[l_idx, ] < 0.01)
  has_snp1 <- snp1 %in% cs_vars
  has_snp2 <- snp2 %in% cs_vars
  cat(sprintf("%s: %d SNP(s), contains SNP%d=%s SNP%d=%s\n",
              cs_name, length(cs_vars),
              snp1, has_snp1, snp2, has_snp2))
  cat(sprintf("  Significant traits (lfsr < 0.01): %s\n",
              paste(sig_traits, collapse = ", ")))
}
```

Two CS are identified,

```{r pip-plot, fig.height=2, fig.width=6}
out <- mvsusie_plot(fit, add_cs = TRUE, lfsr_cutoff = 0.05)
out$pip_plot
```

### Effect plot (mvSuSiE)

The `mvsusie_plot` effect plot shows that each credible set has
effects in only its 5 causal traits, which looks different from the
marginal association plots previously shown.

```{r effect-plot, fig.height=3.5, fig.width=5}
out$effect_plot
```

### Marginal z-scores for CS variants

For comparison, we plot the marginal z-scores for the same
variants that appear in the mvSuSiE credible sets.

```{r z-scores-cs, fig.height=3.5, fig.width=5}
# Extract CS SNPs and their z-scores
cs_z_dat <- data.frame()
for (cs_name in names(fit$sets$cs)) {
  cs_vars <- fit$sets$cs[[cs_name]]
  for (j in cs_vars) {
    df <- data.frame(
      trait = paste0("T", 1:r),
      zscore = Z[j, ],
      pval = pval[j, ],
      snp_cs = paste0("SNP", j, "(", cs_name, ")"),
      cs = cs_name,
      stringsAsFactors = FALSE
    )
    cs_z_dat <- rbind(cs_z_dat, df)
  }
}
cs_z_dat$trait <- factor(cs_z_dat$trait, levels = paste0("T", 1:r))
cs_z_dat$direction <- factor(ifelse(cs_z_dat$zscore > 0, "+1", "-1"),
                             levels = c("-1", "+1"))
cs_z_dat$logp <- -log10(cs_z_dat$pval)

ggplot(cs_z_dat, aes(x = snp_cs, y = trait,
                     fill = direction, size = logp)) +
  geom_point(shape = 21, stroke = 0.5, color = "white") +
  scale_fill_manual(values = c("darkblue", "red"), drop = FALSE) +
  scale_size(range = c(1, 6),
             breaks = unname(round(quantile(cs_z_dat$logp,
                                            seq(0, 1, length.out = 4)), 1))) +
  labs(x = "", y = "", fill = "direction",
       size = expression(-log[10](p)),
       title = "Marginal z-scores") +
  guides(fill = guide_legend(override.aes = list(size = 2)),
         size = guide_legend(override.aes = list(shape = 21, fill = "black",
                                                  color = "white",
                                                  stroke = 0.5))) +
  theme_cowplot(font_size = 10) +
  background_grid(major = "xy", minor = "none",
                  size.major = 0.3, colour.major = "grey85") +
  panel_border(colour = "grey70") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Comparison with univariate fine-mapping

Here we analyze 10 traits seperately using `susie()`:

```{r univariate-comparison}
for (t in c(1:10)) {
  fit_uv <- susie(X, Y[, t], L = 10, standardize = TRUE)
  if (length(fit_uv$sets$cs) > 0) {
    cs1 <- fit_uv$sets$cs[[1]]
    cat(sprintf("Trait %d CS: %d SNPs, contains SNP%d=%s, SNP%d=%s\n",
                t, length(cs1),
                snp1, snp1 %in% cs1,
                snp2, snp2 %in% cs1))
  } else {
    cat(sprintf("Trait %d: no CS found\n", t))
  }
}
```

## The challenge of weak LD contrast

When the effects are smaller relative to noise, both univariate
and multivariate methods lose power. With PVE around 2--3%,
mvSuSiE may find fewer credible sets or may not fully resolve
the two signals.

```{r weak-effects}
set.seed(2)
B_weak <- matrix(0, p, r)
B_weak[snp1, 1:5] <- 1.0
B_weak[snp2, 6:10] <- 1.0
Y_weak <- X %*% B_weak + matrix(rnorm(n * r), n, r)

pve_weak <- sapply(1:r, function(t) {
  var(X %*% B_weak[, t]) / var(Y_weak[, t])
})
cat(sprintf("Per-trait PVE range: %.1f%% -- %.1f%%\n",
            100 * min(pve_weak), 100 * max(pve_weak)))

fit_weak <- mvsusie(X, Y_weak, L = 10,
                    prior_variance = prior)
out <- mvsusie_plot(fit_weak, add_cs = TRUE, lfsr_cutoff = 0.05)
out$pip_plot
out$effect_plot
```