---
title: Handling missing data in mvSuSiE
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling missing data in mvSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

mvSuSiE can handle missing values in the phenotype matrix $Y$. This
is useful in two common scenarios:

1. **Random missing data** (MCAR): individual measurements are
   randomly missing across samples and traits.

2. **Block missing data**: entire traits are unobserved for subsets
   of samples. This arises naturally in cross-population or
   cross-ancestry analyses where different cohorts measure different
   traits.

The default method `missing_y_method = "approximate"` as described in [Zou's thesis (2021)](https://stephenslab.uchicago.edu/assets/papers/yuxin-thesis.pdf) handles both
scenarios. An alternative `"impute"` method, as discussed in [Zou et al (2026)](https://www.nature.com/articles/s41588-025-02486-7) is available and may perform well for 
MCAR with lower missing values, but loses power for block-structured missingness (see below).

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

## Simulate complete data

```{r simulate-data}
data(simdata)
X <- simdata$raw$X
n <- nrow(X); p <- ncol(X); r <- 10

causal <- sort(sample(p, 4))
B <- matrix(0, p, r)
for (j in causal) B[j, ] <- rnorm(r, 0, 0.5)
Y_complete <- X %*% B + matrix(rnorm(n * r), n, r)
cat(sprintf("Complete data: %d x %d x %d (N x J x R)\n", n, p, r))
cat("True causal SNPs:", causal, "\n")
```

Fit the model on complete data for reference.

```{r fit-complete}
prior <- create_mixture_prior(R = r)
fit_complete <- mvsusie(X, Y_complete, L = 10,
                        prior_variance = prior)
cat("Complete data: CS =", length(fit_complete$sets$cs), "\n")
```

We evaluate performance at the credible set (CS) level: a true
positive is a CS that contains at least one causal SNP; a false
positive is a CS containing no causal SNP.

```{r cs-eval-helper}
check_cs <- function(fit, causal, label) {
  cs_list <- fit$sets$cs
  n_cs <- if (is.null(cs_list)) 0 else length(cs_list)
  tp <- 0; fp <- 0
  if (n_cs > 0) {
    for (i in seq_along(cs_list)) {
      if (any(cs_list[[i]] %in% causal)) tp <- tp + 1
      else fp <- fp + 1
    }
  }
  captured <- sapply(causal, function(c) {
    any(sapply(cs_list, function(cs) c %in% cs))
  })
  cat(sprintf("%s: %d CS (%d true, %d false), %d/%d causal captured\n",
              label, n_cs, tp, fp, sum(captured), length(causal)))
}
check_cs(fit_complete, causal, "Complete")
```

## Random missing data (MCAR)

We introduce missing values completely at random at three rates:
5%, 20%, and 50%.

```{r random-missing}
insert_random_na <- function(Y, frac) {
  Y_miss <- Y
  n_miss <- round(frac * length(Y))
  idx <- sample(length(Y), n_miss)
  Y_miss[idx] <- NA
  Y_miss
}

Y_miss_05 <- insert_random_na(Y_complete, 0.05)
Y_miss_20 <- insert_random_na(Y_complete, 0.20)
Y_miss_50 <- insert_random_na(Y_complete, 0.50)
cat("Missing values: 5% =", sum(is.na(Y_miss_05)),
    ", 20% =", sum(is.na(Y_miss_20)),
    ", 50% =", sum(is.na(Y_miss_50)), "\n")
```

### Default method (approximate)

```{r fit-random-approx}
fit_05 <- mvsusie(X, Y_miss_05, L = 10, prior_variance = prior)
fit_20 <- mvsusie(X, Y_miss_20, L = 10, prior_variance = prior)
fit_50 <- mvsusie(X, Y_miss_50, L = 10, prior_variance = prior)
check_cs(fit_05, causal, "5% missing (approx)")
check_cs(fit_20, causal, "20% missing (approx)")
check_cs(fit_50, causal, "50% missing (approx)")
```

```{r compare-random-approx, fig.height=3, fig.width=7}
pdat <- data.frame(
  complete = rep(fit_complete$pip, 3),
  missing = c(fit_05$pip, fit_20$pip, fit_50$pip),
  frac = rep(c("5% missing", "20% missing", "50% missing"), each = p)
)
ggplot(pdat, aes(x = complete, y = missing)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  facet_wrap(~frac) +
  labs(x = "PIP (complete data)", y = "PIP (approximate)") +
  theme_cowplot(font_size = 10)
```

With moderate amounts of missing data, the default `"approximate"`
method gives results largely similar to the complete data analysis,
although some signs of inflation in PIP can be observed in this experiments.

### Impute method

The `"impute"` method also handles MCAR data well:

```{r fit-random-impute}
fit_imp_05 <- mvsusie(X, Y_miss_05, L = 10, prior_variance = prior,
                      missing_y_method = "impute")
fit_imp_20 <- mvsusie(X, Y_miss_20, L = 10, prior_variance = prior,
                      missing_y_method = "impute")
fit_imp_50 <- mvsusie(X, Y_miss_50, L = 10, prior_variance = prior,
                      missing_y_method = "impute")
check_cs(fit_imp_05, causal, "5% missing (impute)")
check_cs(fit_imp_20, causal, "20% missing (impute)")
check_cs(fit_imp_50, causal, "50% missing (impute)")
```

```{r compare-random-impute, fig.height=3, fig.width=7}
pdat <- data.frame(
  complete = rep(fit_complete$pip, 3),
  missing = c(fit_imp_05$pip, fit_imp_20$pip, fit_imp_50$pip),
  frac = rep(c("5% missing", "20% missing", "50% missing"), each = p)
)
ggplot(pdat, aes(x = complete, y = missing)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  facet_wrap(~frac) +
  labs(x = "PIP (complete data)", y = "PIP (impute)") +
  theme_cowplot(font_size = 10)
```

It performs well in terms of FDR but there is power loss particularly as missing rate goes higher.

## Block missing data (cross-population scenario)

In a cross-population analysis, different cohorts may measure
different subsets of traits. We simulate this by making the first
5 traits observed only in the first half of samples, and the last
5 traits only in the second half.

```{r block-missing}
Y_block <- Y_complete
half <- n %/% 2
Y_block[1:half, 6:10] <- NA
Y_block[(half + 1):n, 1:5] <- NA
cat("Block missing pattern:\n")
cat("  Samples 1-", half, ": traits 1-5 observed, 6-10 missing\n")
cat("  Samples", half + 1, "-", n, ": traits 6-10 observed, 1-5 missing\n")
cat("  Total missing:", sum(is.na(Y_block)), "/", length(Y_block),
    sprintf("(%.0f%%)\n", 100 * mean(is.na(Y_block))))
```

The default `"approximate"` method handles this well:

```{r fit-block}
fit_block <- mvsusie(X, Y_block, L = 10,
                     prior_variance = prior)
check_cs(fit_block, causal, "Block (approximate)")
```

```{r compare-block, fig.height=3, fig.width=4}
pdat <- data.frame(complete = fit_complete$pip,
                   block = fit_block$pip)
ggplot(pdat, aes(x = complete, y = block)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  labs(x = "PIP (complete data)", y = "PIP (block missing)") +
  theme_cowplot(font_size = 12)
```

mvSuSiE borrows information across the observed portions to identify
causal signals even when no single sample has all traits observed.

### Why not use impute for block missing?

The `"impute"` method struggles with block-structured missingness
because traits from different blocks are never observed together,
making it impossible to impute correctly:

```{r fit-block-impute}
fit_block_imp <- mvsusie(X, Y_block, L = 10,
                         prior_variance = prior,
                         missing_y_method = "impute")
check_cs(fit_block_imp, causal, "Block (impute)")
```

```{r compare-block-methods, fig.height=3, fig.width=7}
pdat <- data.frame(
  complete = rep(fit_complete$pip, 2),
  missing = c(fit_block$pip, fit_block_imp$pip),
  method = rep(c("approximate (default)", "impute"), each = p)
)
ggplot(pdat, aes(x = complete, y = missing)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  facet_wrap(~method) +
  labs(x = "PIP (complete data)", y = "PIP (block missing)") +
  theme_cowplot(font_size = 10)
```

The `"approximate"` method recovers complete-data PIPs well, while
`"impute"` loses substantial power. 

## Missing data methods

mvSuSiE provides three methods for handling missing data, controlled
by the `missing_y_method` argument:

- `"approximate"` (default): per-outcome centering with per-pattern
  residual variance correction. Fast and accurate; designed for the
  block-structured missingness common in multi-population analyses.

- `"exact"`: uses full residual covariance correction. Produces
  nearly identical results to `"approximate"` (typical PIP
  differences $< 10^{-4}$) at higher computational cost.

- `"impute"`: variational imputation that iteratively estimates
  missing values. Works well for scattered (MCAR) missing data
  at low missing rate but loses power for high or block-structured missingness.

```{r method-comparison}
fit_approx <- mvsusie(X, Y_miss_20, L = 10, prior_variance = prior,
                      estimate_prior_variance = TRUE,
                      missing_y_method = "approximate",
                      standardize = TRUE, verbose = FALSE)
fit_exact <- mvsusie(X, Y_miss_20, L = 10, prior_variance = prior,
                     estimate_prior_variance = TRUE,
                     missing_y_method = "exact",
                     standardize = TRUE, verbose = FALSE)

cat("20% MCAR:\n")
check_cs(fit_approx, causal, "  approximate")
check_cs(fit_exact, causal, "  exact")
cat("  PIP correlation (approx vs exact):", round(cor(fit_approx$pip, fit_exact$pip), 6), "\n")
```

For most applications, the default `"approximate"` method provides
a good balance of speed and accuracy.