---
title: Handling missing data in mvSuSiE
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling missing data in mvSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

mvSuSiE can handle missing values in the phenotype matrix $Y$. This
is useful in two common scenarios:

1. **Random missing data** (MCAR): individual measurements are
   randomly missing across samples and traits.

2. **Block missing data**: entire traits are unobserved for subsets
   of samples. This arises naturally in cross-population or
   cross-ancestry analyses where different cohorts measure different
   traits.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

## Simulate complete data
```{r simulate-data}
data(N3finemapping)
X <- N3finemapping$X
n <- nrow(X); p <- ncol(X); r <- 10

causal <- sort(sample(p, 4))
B <- matrix(0, p, r)
for (j in causal) B[j, ] <- rnorm(r, 0, 0.5)
Y_complete <- X %*% B + matrix(rnorm(n * r), n, r)
cat(sprintf("Complete data: %d x %d x %d (N x J x R)\n", n, p, r))
```

Fit the model on complete data for reference.

```{r fit-complete}
prior <- create_mixture_prior(R = r)
fit_complete <- mvsusie(X, Y_complete, L = 10,
                        prior_variance = prior,
                        estimate_prior_variance = TRUE,
                        standardize = TRUE, tol = 0.01,
                        verbosity = 0)
cat("Complete data: CS =", length(fit_complete$sets$cs), "\n")
```

## Random missing data (MCAR)
We introduce missing values completely at random at three rates:
5%, 20%, and 50%.

```{r random-missing}
insert_random_na <- function(Y, frac) {
  Y_miss <- Y
  n_miss <- round(frac * length(Y))
  idx <- sample(length(Y), n_miss)
  Y_miss[idx] <- NA
  Y_miss
}

Y_miss_05 <- insert_random_na(Y_complete, 0.05)
Y_miss_20 <- insert_random_na(Y_complete, 0.20)
Y_miss_50 <- insert_random_na(Y_complete, 0.50)
cat("Missing values: 5% =", sum(is.na(Y_miss_05)),
    ", 20% =", sum(is.na(Y_miss_20)),
    ", 50% =", sum(is.na(Y_miss_50)), "\n")
```

```{r fit-random-missing}
fit_05 <- mvsusie(X, Y_miss_05, L = 10, prior_variance = prior,
                  estimate_prior_variance = TRUE,
                  standardize = TRUE, tol = 0.01, verbosity = 0)
fit_20 <- mvsusie(X, Y_miss_20, L = 10, prior_variance = prior,
                  estimate_prior_variance = TRUE,
                  standardize = TRUE, tol = 0.01, verbosity = 0)
fit_50 <- mvsusie(X, Y_miss_50, L = 10, prior_variance = prior,
                  estimate_prior_variance = TRUE,
                  standardize = TRUE, tol = 0.01, verbosity = 0)
cat("5% missing: CS =", length(fit_05$sets$cs), "\n")
cat("20% missing: CS =", length(fit_20$sets$cs), "\n")
cat("50% missing: CS =", length(fit_50$sets$cs), "\n")
```

```{r compare-random, fig.height=3, fig.width=7}
pdat <- data.frame(
  complete = rep(fit_complete$pip, 3),
  missing = c(fit_05$pip, fit_20$pip, fit_50$pip),
  frac = rep(c("5% missing", "20% missing", "50% missing"), each = p)
)
ggplot(pdat, aes(x = complete, y = missing)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  facet_wrap(~frac) +
  labs(x = "PIP (complete data)", y = "PIP (missing data)") +
  theme_cowplot(font_size = 10)
```

With moderate amounts of missing data (up to 20%), the results are
very similar to the complete data analysis. Even at 50% missing,
the strongest signals are preserved.

## Block missing data (cross-population scenario)
In a cross-population analysis, different cohorts may measure
different subsets of traits. We simulate this by making the first
5 traits missing for half the samples, and the last 5 traits missing
for the other half. This mimics two cohorts that each measure 5 of
10 traits, plus 5 shared traits are available.

Actually, a cleaner block-missing pattern: traits 1--5 are observed
only in the first half of samples, traits 6--10 only in the second
half. No trait is observed in all samples.

```{r block-missing}
Y_block <- Y_complete
half <- n %/% 2
# First half of samples: traits 6-10 are missing
Y_block[1:half, 6:10] <- NA
# Second half of samples: traits 1-5 are missing
Y_block[(half + 1):n, 1:5] <- NA
cat("Block missing pattern:\n")
cat("  Samples 1-", half, ": traits 1-5 observed, 6-10 missing\n")
cat("  Samples", half + 1, "-", n, ": traits 6-10 observed, 1-5 missing\n")
cat("  Total missing:", sum(is.na(Y_block)), "/", length(Y_block),
    sprintf("(%.0f%%)\n", 100 * mean(is.na(Y_block))))
```

```{r fit-block}
fit_block <- mvsusie(X, Y_block, L = 10,
                     prior_variance = prior,
                     estimate_prior_variance = TRUE,
                     standardize = TRUE, tol = 0.01,
                     verbosity = 0)
cat("Block missing: CS =", length(fit_block$sets$cs), "\n")
```

```{r compare-block, fig.height=3, fig.width=4}
pdat <- data.frame(complete = fit_complete$pip,
                   block = fit_block$pip)
ggplot(pdat, aes(x = complete, y = block)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  labs(x = "PIP (complete data)", y = "PIP (block missing)") +
  theme_cowplot(font_size = 12)
```

mvSuSiE can borrow information across the observed portions to
identify causal signals even when no single sample has all traits
observed. Note that the prior does not model realistic
cross-population assumptions (such as different effect sizes across
ancestries), but the method works given a reasonable prior structure.

## Missing data methods
mvSuSiE provides three methods for handling missing data, controlled
by the `missing_y_method` argument:

- `"approximate"` (default): per-outcome centering with per-pattern
  residual variance correction. Fast and accurate when the residual
  covariance is approximately diagonal.

- `"exact"`: uses full residual covariance correction. Guarantees
  correct ELBO but slower.

- `"impute"`: variational imputation that iteratively estimates
  missing values. Does not guarantee monotone ELBO.

```{r missing-methods}
fit_approx <- mvsusie(X, Y_miss_20, L = 10, prior_variance = prior,
                      estimate_prior_variance = TRUE,
                      missing_y_method = "approximate",
                      standardize = TRUE, tol = 0.01, verbosity = 0)
fit_exact <- mvsusie(X, Y_miss_20, L = 10, prior_variance = prior,
                     estimate_prior_variance = TRUE,
                     missing_y_method = "exact",
                     standardize = TRUE, tol = 0.01, verbosity = 0)

cat("Approximate method: CS =", length(fit_approx$sets$cs), "\n")
cat("Exact method: CS =", length(fit_exact$sets$cs), "\n")
cat("PIP correlation:", cor(fit_approx$pip, fit_exact$pip), "\n")
```

For most applications, the default `"approximate"` method provides
a good balance of speed and accuracy.
