---
title: Introduction to mvSuSiE for multi-trait fine-mapping
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to mvSuSiE for multi-trait fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette walks through the steps of a basic mvSuSiE multi-trait
fine-mapping analysis using individual-level and summary data.

For detailed interpretation of the output including PIP, credible sets,
LFSR, outcome-specific PIPs, and per-effect coefficient
estimates, see the [Interpreting and visualizing mvSuSiE
results](visualization_and_interpretation.html) vignette.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

First, load the necessary packages, and to ensure reproducibility set
the seed.

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

Next, load the data set we will analyze.

```{r load-data}
data(simdata)
```

The data consist of an $N \times J$ genotype matrix `X` and an $N
\times R$ phenotype matrix `Y` where $N=574$ is the number of samples, $J=1001$
is the number of SNPs, and $R=20$ is the number of traits/phenotypes.

```{r inspect-data}
X <- simdata$raw$X
Y <- simdata$raw$Y
dim(X)
dim(Y)
```

## Specify the prior

Some care should be taken to choose the prior for your mvSuSiE
analysis. To simplify the example we have defined the prior for you:

```{r create-prior}
prior <- create_mixture_prior(list(matrices = simdata$par$U,
                                   weights = simdata$par$w),
                              null_weight = 0)
```

More generally, we recommend taking a "data-driven" approach to
estimating the weights and the covariance matrices in the prior. For
example, when there are many candidate fine-mapping regions in your
study, you could fit the prior to all the top associations from each
of the candidate fine-mapping regions. This data-driven approach is
quite similar to the approach used in [mashr][mashr], and we recommend
looking through the vignettes in the mashr package for further guidance.
See also the [prior specification](prior_specification.html) vignette.

## Fit the mvSuSiE model

Now that we have set up the prior, we are ready to run mvsusie:

```{r run-mvsusie}
fit <- mvsusie(X,Y,standardize = TRUE,
               prior_variance = prior,
               residual_variance = simdata$par$V,
               estimate_prior_variance = TRUE)
```

In this call, we have provided mvSuSiE with the four key pieces of
information needed to run the analysis: (1) the genotype matrix; (2)
the phenotype matrix; (3) the prior on the effects; and (4) the
residual variance-covariance matrix (an $R \times R$ matrix).

The model identified the following credible sets:

```{r credible-sets}
cat("Credible sets:", length(fit$sets$cs), "\n")
fit$sets$cs
```

## mvSuSiE with summary data

mvsusieR also provides a "summary data" interface, building on the
ideas developed in the [susie-rss paper][susie-rss]. Different
combinations of summary statistics are accepted, but here we will
focus on the most common case in which the *z*-scores from the
association tests are used. These *z*-scores should be stored in a $J
\times R$ matrix:

```{r summary-data-1}
Z <- simdata$sumstats$z
dim(Z)
```

In addition to the *z*-scores, the sample size and $R \times R$ LD
(correlation) matrix are needed. (Strictly speaking, the sample size
is not required, but the estimates may be less accurate if the sample
size is not provided.)

```{r summary-data-2}
n <- simdata$sumstats$n
R <- simdata$sumstats$LD
```

Now we can fit an mvSuSiE model to these summary data via the
`mvsusie_rss` interface:

```{r mvsusie-rss}
fit_rss <- mvsusie_rss(Z,R,n,
                       prior_variance = prior,
                       residual_variance = simdata$par$V,
                       estimate_prior_variance = TRUE)
```

## Comparing individual-level and summary data results

When the genotypes are standardized in `mvsusie` (`standardize =
TRUE`) and the LD estimates are obtained from the same data that were
used to compute the *z*-scores, both `mvsusie` and `mvsusie_rss`
give similar PIPs:

```{r raw-vs-summary, fig.height=3, fig.width=3.5}
ggplot(data.frame(pip = fit$pip, pip_rss = fit_rss$pip),
       aes(x = pip, y = pip_rss)) +
  geom_point(shape = 20, size = 2, color = "royalblue") +
  geom_abline(intercept = 0, slope = 1, color = "black", linetype = "dotted") +
  labs(x = "mvsusie PIP", y = "mvsusie_rss PIP") +
  theme_cowplot(font_size = 12)
```

The PIPs are close but not identical because `mvsusie()` defaults to
`estimate_residual_variance = TRUE` (re-estimating the residual
variance from the data), whereas `mvsusie_rss()` defaults to `FALSE`
(since the residual variance is difficult to estimate reliably from
summary statistics alone, as noted in [Zou *et al.*][susie-rss]).
For a detailed comparison of all input formats and the effect of
estimating the residual variance, see the
[input data formats](input_data_format.html) vignette.

Of course, the results are also not always expected to be the same
when the LD is estimated from an external reference panel.

## Estimating the residual variance

In the above examples, we set the residual variance to the value that
was used to simulate the data. In practice of course we will not have
access to this so we will have to estimate the residual variance somehow.

When one has access to the summary data (specifically, the z-scores)
for a large pool of independent SNPs that are expected to have little
to no effect on the traits, one can use these z-scores to estimate the
residual variance. Here, in this example data set, we do not have
access to a large pool of independent "null SNPs", but we can still
use the data we have available to obtain a rough estimate of the
residual variance.  Specifically, we take as the "null SNPs" the SNPs
in which the z-scores are all less than 2 in magnitude:

```{r get-null-markers}
Z <- simdata$sumstats$z
null_markers <- which(apply(abs(Z),1,max) < 2)
length(null_markers)
Z <- Z[null_markers,]
Vest <- cov(Z)
```

The estimate of the residual covariance matrix using these SNPs is
reasonable, although presumably one could do better if we had the
z-scores for more SNPs that were not in LD with the causal SNPs:

```{r compare-residual-estimate, fig.height=2.75, fig.width=3}
pdat <- data.frame(true = as.vector(simdata$par$V) + rnorm(400)/30,
                   estimated = as.vector(Vest))
ggplot(pdat,aes(x = true,y = estimated)) +
  geom_point(shape = 20,size = 2.5,color = "royalblue") +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  theme_cowplot(font_size = 12)
```

[mashr]: https://github.com/stephenslab/mashr
[susie-rss]: https://doi.org/10.1371/journal.pgen.1010299
