---
title: Introduction to mvsusie for multi-trait fine-mapping
author: Yuxin Zou, Gao Wang, Peter Carbonetto and Matthew Stephens
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to mvsusie for multi-trait fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The aim of this vignette is to work through the steps of a basic
mvSuSiE multi-trait fine-mapping analysis. We analyze a simulated
fine-mapping data set with realistic genotypes.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

First, load the necessary packages, and to ensure reproducibility set
the seed.

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

Next, load the data set we will analyze.

```{r load-data}
data(simdata)
```

The data consist of an $N \times J$ genotype matrix `X` and an $N
\times R$ phenotype matrix `Y` where $N$ is the number of samples, $J$
is the number of SNPs, and $R$ is the number of traits/phenotypes.

```{r inspect-data}
X <- simdata$raw$X
Y <- simdata$raw$Y
dim(X)
dim(Y)
```

Specify the prior
-----------------

Some care should be taken to choose the prior for your mvSuSiE
analysis. To simplify the example we have defined the prior for you:

```{r create-prior}
prior <- create_mixture_prior(list(matrices = simdata$par$U,
                                   weights = simdata$par$w),
                              null_weight = 0)
```
							  
More generally, we recommend taking a "data-driven" approach to
estimating the weights and the covariance matrices in the prior. For
example, when there are many candidate fine-mapping regions in your
study, you could fit the prior to all the top associations from each
of the candidate fine-mapping regions. This data-driven approach is
quite similar to the approach used in [mashr][mashr], and we recommend
looking through the vignettes in the mashr package for further guidance.

Fit the mvSuSiE model
---------------------

Now that we have set up the prior, we are ready to run mvsusie:

```{r run-mvsusie}
fit <- mvsusie(X,Y,standardize = TRUE,
               prior_variance = prior,
               residual_variance = simdata$par$V,
			   estimate_prior_variance = TRUE,
			   tol = 0.01)
```

In this call, we have provided mvSuSiE with the four key pieces of
information need to run the analysis: (1) the genotype matrix; (2) the
phenotype matrix, (3) the prior on the effects; and (4) the residual
variance-covariance matrix (an $R \times R$ matrix).

The "tol" parameter tells the IBSS model fitting algorithm when to stop
iterating; for a final analysis, you may want to set a more stringent
tolerance (a number closer to zero).

Identifying the causal SNPs: Credible Sets 
------------------------------------------

By default, the "mvsusie" function fits a model with L = 10 single
effects. mvSuSiE is generally robust to misspecification of L so long
as L is chosen to be larger than the true number of effects. mvSuSiE
prunes unneeded single effects by estimating a scaling coefficient in
the prior; we set `estimate_prior_variance = TRUE` to turn on this
pruning step. Indeed, mvsusie pruned 7 out of 10 of the single effects,
resulting in three 95% credible sets (CSs), each of which contain
exactly one of the true causal SNPs:

```{r credible-sets}
fit$sets$cs
```

In the third CS, mvSuSiE cannot decide which SNP is the causal SNP
because the candidate SNPs are strongly correlated with each other:

```{r cs-purity}
fit$sets$purity
```

Examining the posterior inclusion probabilties (PIPs) for these SNPs,
there is a subset of 3 SNPs that account for >58% of probability for
being causal:

```{r cs-3}
markers <- fit$sets$cs$L3
fit$pip[markers]
```

Additionally, the "sentinel" SNP---the SNP with the highest PIP---as a
PIP of nearly 40%.

This is all summarized visually in a PIP plot:

```{r pip-plot, fig.height=2, fig.width=6}
mvsusie_plot_simple(fit)$pip_plot
```

Identifying which traits are affected
-------------------------------------

In a multi-trait analysis, we are not just interested in identifying
the causal SNPs, but also _which traits_ are affected by these causal
SNPs.

Let's take the first CS as an example. This is an example of a causal
SNP that affects most traits:

```{r, fig.height=2.5, fig.width=3.5}
l <- 1
i <- 335
pdat <- data.frame(bhat = simdata$Btrue[i,],
                   coef = fit$b1_rescaled[l,i+1,],
                   lfsr = fit$single_effect_lfsr[l,])
ggplot(pdat,aes(x = bhat,y = coef,color = log10(lfsr),pch = 20)) +
  geom_point(shape = 20,size = 2.5) +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_color_gradient2(low = "lightskyblue",mid = "gold",high = "orangered",
                        midpoint = -40) +
  labs(x = "true coef",y = "mvsusie estimate") +
  theme_cowplot(font_size = 12)
```

Indeed, the effects of this causal SNP are significant (very low local
false sign rates) for most of the 20 traits. (A local false sign rate
can be thought of as an alternative to a *p*-value, but is typically
slightly more conservative than a *p*-value.)

```{r, fig.height=2.5, fig.width=3.5}
l <- 3
i <- 493
pdat <- data.frame(bhat = simdata$Btrue[i,],
                   coef = fit$b1_rescaled[l,i+1,],
                   lfsr = fit$single_effect_lfsr[l,])
ggplot(pdat,aes(x = bhat,y = coef,color = log10(lfsr),pch = 20)) +
  geom_point(shape = 20,size = 2.5) +
  scale_color_gradient2(low = "lightskyblue",mid = "gold",high = "orangered",
                        midpoint = -3) +
  labs(x = "true coef",y = "mvsusie estimate") +
  theme_cowplot(font_size = 12)
```

mvSuSiE with summary data: mvSuSiE-RSS
--------------------------------------

```{r mvsusie-rss}
n <- simdata$sumstats$n
Z <- simdata$sumstats$z
R <- simdata$sumstats$LD
fit_rss <- mvsusie_rss(Z,R,n,
                       prior_variance = prior,
					   residual_variance = simdata$par$V,
					   estimate_prior_variance = TRUE,
					   tol = 0.01)
```

```{r}
l <- 1
i <- 335
fit$b1[l,i,]
fit_rss$b1[l,i,]
```

[mashr]: https://github.com/stephenslab/mashr
