---
title: Per-trait sample sizes in mvSuSiE RSS
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Per-trait sample sizes in mvSuSiE RSS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

When combining summary statistics from different studies ---
for example, a GWAS trait ($N = 300\text{K}$) and an eQTL trait
($N = 700$) --- each trait has its own sample size.  The z-score-only
interface to `mvsusie_rss()` takes a single scalar $N$ and constructs 
sufficient statistics as
$X^TX = (N-1)\,R$ and $X^TY = \sqrt{N-1}\,Z$. It cannot represent this difference directly.  
Because mvSuSiE RSS assumes a **shared LD
structure** across all traits, which is reasonable when all studies are
from the same ancestry, even if their sample sizes differ.

This vignette shows a workaround converting
z-scores with per-trait sample sizes into `Bhat`/`Shat` inputs that at least will 
encode the per-trait uncertainty into standard error, such that traits with
smaller $N$ get larger standard errors, which correctly reduces
their influence on the posterior.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

We simulate 5 traits on the same genotypes, then assign
different "effective" sample sizes to each trait to mimic the
scenario where z-scores come from studies of different sizes.

```{r simulate-data}
data(simdata)
X <- simdata$raw$X
n <- nrow(X)
p <- ncol(X)
r <- 5
cat(sprintf("Genotype matrix: %d samples x %d SNPs\n", n, p))

causal <- sort(sample(p, 3))
B <- matrix(0, p, r)
for (j in causal) B[j, ] <- rnorm(r, 0, 0.5)
Y <- X %*% B + matrix(rnorm(n * r), n, r)
cat("Causal SNPs:", causal, "\n")

# Per-trait sample sizes: traits 1-2 use full data,
# traits 3-4 are moderate, trait 5 is small (eQTL-like).
N_vec <- c(n, n, round(n / 5), round(n / 5), round(n / 20))
cat("Per-trait N:", N_vec, "\n")
```

```{r setup}
prior <- create_mixture_prior(R = r)
resid_var <- diag(r)
Z <- calc_z(X, Y, center = TRUE, scale = TRUE)
LD <- cor(X)
```

## Convert z-scores to Bhat / Shat

On the standardized scale ($\mathrm{var}(X_j) = 1$,
$\mathrm{var}(Y_r) = 1$), the marginal z-score follows a
$t$-distribution with $N_r - 2$ degrees of freedom.  The exact
conversion is:

$$
\mathrm{Shat}_{jr} = \frac{1}{\sqrt{N_r - 2 + Z_{jr}^2}},
\qquad
\mathrm{Bhat}_{jr} = Z_{jr}\;\mathrm{Shat}_{jr}.
$$

For large $N_r$ and moderate $Z$ this simplifies to
$\mathrm{Shat} \approx 1/\sqrt{N_r}$ and
$\mathrm{Bhat} \approx Z/\sqrt{N_r}$.  The exact form is
preferred when $N_r$ is small or $|Z|$ is large, because it
properly accounts for the finite-sample PVE adjustment.

```{r convert-z}
Bhat <- matrix(0, p, r)
Shat <- matrix(0, p, r)
for (rr in seq_len(r)) {
  denom <- sqrt(N_vec[rr] - 2 + Z[, rr]^2)
  Bhat[, rr] <- Z[, rr] / denom
  Shat[, rr] <- 1 / denom
}

cat("Shat magnitude at a null SNP (SNP 1):\n")
cat("  per trait:", round(Shat[1, ], 4), "\n")
cat("  ratio to max-N trait:", round(Shat[1, ] / Shat[1, 1], 2), "\n")
```

Traits with smaller $N$ get larger standard errors (roughly
$\sqrt{N_{\max}/N_r}$ times larger).

## Fit with per-trait sample sizes

Pass `Bhat`, `Shat`, and a `varY` to `mvsusie_rss()`.
In our simulation on the standardized scale, `varY` is an identity
matrix, but in practice this can be left unspecified so that the
residual correlation is estimated internally from null z-scores.
We use `estimate_residual_variance = FALSE` (the default for
summary statistics).

Notice that `mvsusie_rss()` still requires a scalar `N`.  Inside the function,
`N` is used for the PVE adjustment (cf Zou 2022). We pass
`N = max(N_vec)` so that the large-$N$ traits (which dominate
discovery) get the correct adjustment.  For the small-$N$ traits the
PVE correction is slightly too weak but the primary sample-size information
is already encoded in `Shat`, which correctly inflates the standard
errors:

```{r fit-models}
varY_id <- diag(r)

# Per-trait N via Bhat/Shat conversion
fit_diff <- mvsusie_rss(Bhat = Bhat, Shat = Shat, R = LD,
                         N = max(N_vec), varY = varY_id, L = 10,
                         prior_variance = prior,
                         residual_variance = resid_var,
                         estimate_residual_variance = FALSE,
                         estimate_prior_variance = TRUE)

# Baseline: same N for all traits (ignoring the difference)
Bhat0 <- Shat0 <- matrix(0, p, r)
for (rr in seq_len(r)) {
  denom <- sqrt(n - 2 + Z[, rr]^2)
  Bhat0[, rr] <- Z[, rr] / denom
  Shat0[, rr] <- 1 / denom
}
fit_same <- mvsusie_rss(Bhat = Bhat0, Shat = Shat0, R = LD,
                         N = n, varY = varY_id, L = 10,
                         prior_variance = prior,
                         residual_variance = resid_var,
                         estimate_residual_variance = FALSE,
                         estimate_prior_variance = TRUE)
```

## Compare results

The PIPs and credible sets are similar because variable selection is
driven primarily by the large-$N$ traits.

```{r compare-pip}
cat("Credible sets (same N):", length(fit_same$sets$cs), "\n")
cat("Credible sets (diff N):", length(fit_diff$sets$cs), "\n")
cat("PIP correlation:", round(cor(fit_same$pip, fit_diff$pip), 4), "\n")
```

However, traits with smaller $N$ produce larger (less confident) LFSR values:

```{r compare-lfsr}
trait_labels <- paste0("Trait ", 1:r, " (N=", N_vec, ")")
cat("Single-effect LFSR per credible set:\n\n")
for (i in seq_along(fit_same$sets$cs)) {
  cs_name <- names(fit_same$sets$cs)[i]
  cat(cs_name, ":\n")
  lfsr_s <- fit_same$single_effect_lfsr[i, ]
  lfsr_d <- fit_diff$single_effect_lfsr[i, ]
  for (rr in 1:r) {
    flag <- if (lfsr_d[rr] > lfsr_s[rr] + 0.01) " *" else ""
    cat(sprintf("  %-20s  same=%.4f  diff=%.4f%s\n",
                trait_labels[rr], lfsr_s[rr], lfsr_d[rr], flag))
  }
  cat("\n")
}
```

Entries marked with `*` show traits where the per-trait $N$
correctly increases the LFSR.

```{r lfsr-plot, fig.height=3, fig.width=8}
# Gather LFSR for all effects into a data frame
lfsr_list <- list()
for (i in seq_along(fit_same$sets$cs)) {
  cs_name <- names(fit_same$sets$cs)[i]
  lfsr_list[[length(lfsr_list) + 1]] <- data.frame(
    lfsr_same = fit_same$single_effect_lfsr[i, ],
    lfsr_diff = fit_diff$single_effect_lfsr[i, ],
    trait = trait_labels,
    cs = cs_name, stringsAsFactors = FALSE)
}
lfsr_dat <- do.call(rbind, lfsr_list)
lfsr_dat$trait <- factor(lfsr_dat$trait, levels = trait_labels)

ggplot(lfsr_dat, aes(x = lfsr_same, y = lfsr_diff, color = trait)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "LFSR (same N for all traits)",
       y = "LFSR (per-trait N)",
       color = "") +
  theme_cowplot(font_size = 11) +
  theme(legend.position = "right")
```

Points above the diagonal are traits where using per-trait $N$
makes the model less confident, which is the correct behavior for
traits with smaller sample sizes.