---
title: Prediction with mvSuSiE
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prediction with mvSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates how to extract coefficient estimates and
make predictions from a fitted mvSuSiE model using the standard R
`coef()` and `predict()` methods.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

## Simulate data and split into training and test sets

We simulate multi-trait data using realistic genotypes and split the
samples into training (80%) and test (20%) sets.

```{r simulate-data}
data(simdata)
X <- simdata$raw$X
n <- nrow(X)
p <- ncol(X)
r <- 10

# Simulate sparse effects: 4 causal SNPs affecting all traits
causal <- sort(sample(p, 4))
B <- matrix(0, p, r)
for (j in causal) B[j, ] <- rnorm(r, 0, 0.5)
Y <- X %*% B + matrix(rnorm(n * r), n, r)
cat(sprintf("Data: %d samples, %d SNPs, %d traits\n", n, p, r))
cat("Causal SNPs:", causal, "\n")
```

```{r train-test-split}
train_idx <- sample(n, round(0.8 * n))
test_idx <- setdiff(seq_len(n), train_idx)
X_train <- X[train_idx, ]
Y_train <- Y[train_idx, ]
X_test <- X[test_idx, ]
Y_test <- Y[test_idx, ]
cat(sprintf("Training: %d samples, Test: %d samples\n",
            length(train_idx), length(test_idx)))
```

## Fit mvSuSiE and extract coefficients

We fit mvSuSiE using a canonical mixture prior. This is the simplest
approach and requires no additional setup.

```{r fit-model}
prior <- create_mixture_prior(R = r)
fit <- mvsusie(X_train, Y_train, L = 10,
               prior_variance = prior)
cat("Credible sets:", length(fit$sets$cs), "\n")
```

The `coef()` method returns a $(J+1) \times R$ matrix. The first
row is the intercept; the remaining $J$ rows are the regression
coefficients.

```{r coef-extract}
beta_hat <- coef(fit)
cat(sprintf("Coefficient matrix: %d x %d (including intercept)\n",
            nrow(beta_hat), ncol(beta_hat)))
```

```{r coef-compare, fig.height=3, fig.width=3.5}
beta_est <- beta_hat[-1, ]  # Remove intercept row
pdat <- data.frame(true = as.vector(B),
                   estimated = as.vector(beta_est))
ggplot(pdat, aes(x = true, y = estimated)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  labs(x = "True coefficient", y = "Estimated coefficient") +
  theme_cowplot(font_size = 12)
```

## In-sample and out-of-sample prediction

Calling `predict()` without a new data matrix returns the fitted
values from the training data. To predict on new data, pass the
test genotype matrix via the `newx` argument.

```{r prediction}
Y_fitted <- predict(fit)
Y_pred <- predict(fit, newx = X_test)
cat(sprintf("In-sample RMSE: %.4f\n", sqrt(mean((Y_train - Y_fitted)^2))))
cat(sprintf("Out-of-sample RMSE: %.4f\n", sqrt(mean((Y_test - Y_pred)^2))))
```

```{r prediction-plot, fig.height=3.5, fig.width=7}
pdat <- data.frame(
  observed = c(Y_test[, 1], Y_test[, 2]),
  predicted = c(Y_pred[, 1], Y_pred[, 2]),
  trait = rep(c("Trait 1", "Trait 2"), each = nrow(Y_test))
)
ggplot(pdat, aes(x = observed, y = predicted)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  facet_wrap(~trait) +
  labs(x = "Observed", y = "Predicted") +
  theme_cowplot(font_size = 12)
```

Per-trait prediction accuracy:

```{r per-trait-accuracy}
cor_per_trait <- sapply(seq_len(r), function(i)
  cor(Y_pred[, i], Y_test[, i]))
rmse_per_trait <- sapply(seq_len(r), function(i)
  sqrt(mean((Y_test[, i] - Y_pred[, i])^2)))

results <- data.frame(
  trait = paste0("trait", seq_len(r)),
  correlation = round(cor_per_trait, 3),
  rmse = round(rmse_per_trait, 4)
)
results
```

## Using a data-driven prior for prediction

A data-driven prior learned via [mashr][mashr] can potentially
improve predictions by better capturing the true effect-sharing
patterns across traits. See the
[prior specification](prior_specification.html) vignette for
details on constructing data-driven priors.

```{r data-driven-prior, message=FALSE}
library(mashr)

# Learn prior from marginal z-scores
Z_train <- calc_z(X_train, Y_train, center = TRUE, scale = TRUE)
mash_data <- mash_set_data(Bhat = Z_train,
                           Shat = matrix(1, nrow(Z_train), r))
U_c <- cov_canonical(mash_data)
m_fit <- mash(mash_data, Ulist = U_c, outputlevel = 0)
prior_dd <- create_mixture_prior(fitted_g = m_fit$fitted_g)

# Fit with data-driven prior
fit_dd <- mvsusie(X_train, Y_train, L = 10,
                  prior_variance = prior_dd)
```

```{r compare-priors, fig.height=3, fig.width=4}
Y_pred_dd <- predict(fit_dd, newx = X_test)
cor_dd <- sapply(seq_len(r), function(i) cor(Y_pred_dd[, i], Y_test[, i]))
pdat <- data.frame(canonical = cor_per_trait, data_driven = cor_dd)
ggplot(pdat, aes(x = canonical, y = data_driven)) +
  geom_point(shape = 20, size = 3, color = "royalblue") +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  labs(x = "Correlation (canonical prior)",
       y = "Correlation (data-driven prior)") +
  theme_cowplot(font_size = 12)
```

In this simple simulation, the canonical and data-driven priors
give same prediction accuracy. The data-driven approach is more
likely to help when the true effect-sharing patterns are complex
and poorly captured by canonical patterns.

[mashr]: https://github.com/stephenslab/mashr
