---
title: Prediction with mvSuSiE
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prediction with mvSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates how to extract coefficient estimates and
make predictions from a fitted mvSuSiE model using the standard R
`coef()` and `predict()` methods.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

Simulate data and split into training and test sets
----------------------------------------------------

We simulate multi-trait data using realistic genotypes and split the
samples into training (80%) and test (20%) sets.

```{r simulate-data}
data(N3finemapping)
X <- N3finemapping$X
n <- nrow(X)
p <- ncol(X)
r <- 10

# Simulate sparse effects: 4 causal SNPs affecting all traits
causal <- sort(sample(p, 4))
B <- matrix(0, p, r)
for (j in causal) B[j, ] <- rnorm(r, 0, 0.5)
Y <- X %*% B + matrix(rnorm(n * r), n, r)
cat(sprintf("Data: %d samples, %d SNPs, %d traits\n", n, p, r))
cat("Causal SNPs:", causal, "\n")
```

```{r train-test-split}
train_idx <- sample(n, round(0.8 * n))
test_idx <- setdiff(seq_len(n), train_idx)
X_train <- X[train_idx, ]
Y_train <- Y[train_idx, ]
X_test <- X[test_idx, ]
Y_test <- Y[test_idx, ]
cat(sprintf("Training: %d samples, Test: %d samples\n",
            length(train_idx), length(test_idx)))
```

Fit the model
-------------

```{r fit-model}
prior <- create_mixture_prior(R = r)
fit <- mvsusie(X_train, Y_train, L = 10,
               prior_variance = prior,
               estimate_prior_variance = TRUE,
               standardize = TRUE, tol = 0.01)
cat("Credible sets:", length(fit$sets$cs), "\n")
```

Extracting coefficients
-----------------------

The `coef()` method returns a $(J+1) \times R$ matrix. The first
row is the intercept; the remaining $J$ rows are the regression
coefficients.

```{r coef-extract}
beta_hat <- coef(fit)
cat(sprintf("Coefficient matrix: %d x %d (including intercept)\n",
            nrow(beta_hat), ncol(beta_hat)))
cat("Intercept:", round(beta_hat[1, ], 3), "\n")
```

We can compare the estimated coefficients to the true values.

```{r coef-compare, fig.height=3, fig.width=3.5}
beta_est <- beta_hat[-1, ]  # Remove intercept row
pdat <- data.frame(true = as.vector(B),
                   estimated = as.vector(beta_est))
ggplot(pdat, aes(x = true, y = estimated)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  labs(x = "True coefficient", y = "Estimated coefficient") +
  theme_cowplot(font_size = 12)
```

In-sample prediction
--------------------

Calling `predict()` without a new data matrix returns the fitted
values from the training data.

```{r in-sample}
Y_fitted <- predict(fit)
cat(sprintf("Fitted values: %d x %d\n", nrow(Y_fitted), ncol(Y_fitted)))
rmse_train <- sqrt(mean((Y_train - Y_fitted)^2))
cat(sprintf("In-sample RMSE: %.4f\n", rmse_train))
```

Out-of-sample prediction
-------------------------

To predict on new data, pass the test genotype matrix via the
`newx` argument.

```{r out-of-sample}
Y_pred <- predict(fit, newx = X_test)
cat(sprintf("Predictions: %d x %d\n", nrow(Y_pred), ncol(Y_pred)))
rmse_test <- sqrt(mean((Y_test - Y_pred)^2))
cat(sprintf("Out-of-sample RMSE: %.4f\n", rmse_test))
```

We can also compute per-trait RMSE to see which traits are predicted
better than others.

```{r per-trait-rmse}
rmse_per_trait <- sapply(seq_len(r), function(i)
  sqrt(mean((Y_test[, i] - Y_pred[, i])^2)))
names(rmse_per_trait) <- paste0("trait", seq_len(r))
round(rmse_per_trait, 4)
```

```{r prediction-plot, fig.height=3, fig.width=7}
# Show predictions for first two traits
pdat <- data.frame(
  observed = c(Y_test[, 1], Y_test[, 2]),
  predicted = c(Y_pred[, 1], Y_pred[, 2]),
  trait = rep(c("Trait 1", "Trait 2"), each = nrow(Y_test))
)
ggplot(pdat, aes(x = observed, y = predicted)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  facet_wrap(~trait) +
  labs(x = "Observed", y = "Predicted") +
  theme_cowplot(font_size = 12)
```

Prediction with univariate response
------------------------------------

The `predict()` and `coef()` methods also work for single-trait
models fitted with `mvsusie()`.

```{r univariate-prediction}
Y1 <- Y_train[, 1]
fit1 <- mvsusie(X_train, Y1, L = 10, standardize = TRUE, tol = 0.01)
Y1_pred <- predict(fit1, newx = X_test)
cat(sprintf("Single-trait test RMSE: %.4f\n",
            sqrt(mean((Y_test[, 1] - Y1_pred)^2))))
```
