---
title: Specifying and estimating priors in mvSuSiE
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specifying and estimating priors in mvSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette covers the choice of prior on effect sizes for mvSuSiE analysis from
simplest naive approach (a fixed diagonal prior) to more flexible priors.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

## Simulate data

We use realistic genotypes and simulate 10 correlated traits with
4 causal SNPs.

```{r simulate-data}
data(simdata)
X <- simdata$raw$X
n <- nrow(X); p <- ncol(X); r <- 10

causal <- sort(sample(p, 4))
B <- matrix(0, p, r)
for (j in causal) B[j, ] <- rnorm(r, 0, 0.5)
Y <- X %*% B + matrix(rnorm(n * r), n, r)
cat(sprintf("Data: %d samples, %d SNPs, %d traits\n", n, p, r))
```

## Approach 1: Fixed diagonal prior

The simplest prior for getting started is a diagonal matrix scaled
to represent the expected heritability per trait. Setting
$U = 0.2 \times \text{diag}(\text{var}(Y))$ as initial value corresponds to
approximately 20% heritability per trait under independent effects, assuming $Y$ is standardized. 
This 20% will be updated by default with `estimate_prior_variance = TRUE` but the 
prior structure, the $\text{diag}(\text{var}(Y))$ part, will stay the same.

```{r diagonal-prior}
Y_var <- apply(Y, 2, var)
U_diag <- 0.2 * diag(Y_var)
prior_diag <- create_mixture_prior(
  list(matrices = list(U_diag), weights = 1)
)
cat("Prior type: single diagonal matrix\n")
cat("Number of components:", prior_diag$n_component, "\n")
```

```{r fit-diagonal}
fit_diag <- mvsusie(X, Y, L = 10,
                    prior_variance = prior_diag,
                    estimate_prior_variance = TRUE,
                    standardize = TRUE)
cat("Credible sets:", length(fit_diag$sets$cs), "\n")
```

## Approach 2: Canonical mixture prior

A more flexible approach uses `create_mixture_prior(R = r)` which
automatically generates a set of canonical covariance matrices. These
include singleton matrices (effect in one trait only) and shared
patterns with varying correlations.

```{r canonical-prior}
prior_canon <- create_mixture_prior(R = r)
cat("Components:", prior_canon$n_component, "\n")
```

```{r fit-canonical}
fit_canon <- mvsusie(X, Y, L = 10,
                     prior_variance = prior_canon,
                     estimate_prior_variance = TRUE,
                     standardize = TRUE)
cat("Credible sets:", length(fit_canon$sets$cs), "\n")
```

## Approach 3: Data-driven prior from mashr

For large-scale studies with many candidate regions, the recommended
approach is to learn the prior from the data using
[mashr](https://github.com/stephenslab/mashr). The idea is to fit
mashr to the top associations from each region, then use the fitted
prior in mvSuSiE.

```{r mashr-prior}
# Compute z-scores for mashr
Z <- calc_z(X, Y, center = TRUE, scale = TRUE)

# Set up mashr data
mash_data <- mashr::mash_set_data(Bhat = Z,
                                  Shat = matrix(1, p, r))

# Fit mashr (simplified for this example)
U_c <- mashr::cov_canonical(mash_data)
m_fit <- mashr::mash(mash_data, Ulist = U_c, outputlevel = 0)

# Create prior from mashr output
prior_mash <- create_mixture_prior(fitted_g = m_fit$fitted_g)
cat("Components from mashr:", prior_mash$n_component, "\n")
```

```{r fit-mashr}
fit_mash <- mvsusie(X, Y, L = 10,
                    prior_variance = prior_mash,
                    estimate_prior_variance = TRUE,
                    standardize = TRUE)
cat("Credible sets:", length(fit_mash$sets$cs), "\n")
```

## Comparing priors

```{r compare-priors, fig.height=3, fig.width=7}
pdat <- data.frame(
  pip = c(fit_diag$pip, fit_canon$pip, fit_mash$pip),
  prior = rep(c("Diagonal", "Canonical", "mashr"),
              each = p),
  pos = rep(seq_len(p), 3)
)
ggplot(pdat, aes(x = pos, y = pip)) +
  geom_point(shape = 20, size = 1, color = "royalblue", alpha = 0.5) +
  facet_wrap(~prior) +
  geom_vline(xintercept = causal, linetype = "dashed",
             color = "red", alpha = 0.5) +
  labs(x = "SNP position", y = "PIP") +
  theme_cowplot(font_size = 10)
```

All three priors identify the causal variants (red dashed lines) for this trivial example, but
in practice the mixture priors can provide better trait-specific resolution, as shown in 
[Zou et al (2026)](https://www.nature.com/articles/s41588-025-02486-7).

## Mixture weight estimation

When using a mixture prior, the `estimate_prior_mixture_weights`
option (default `TRUE`) allows the relative weights of the mixture
components to be estimated from the data. This is useful for
pruning irrelevant components and adapting to the data.

```{r mixture-weights}
fit_fixed_w <- mvsusie(X, Y, L = 10,
                       prior_variance = prior_canon,
                       estimate_prior_variance = TRUE,
                       estimate_prior_mixture_weights = FALSE,
                       standardize = TRUE, verbose = FALSE)
fit_est_w <- mvsusie(X, Y, L = 10,
                     prior_variance = prior_canon,
                     estimate_prior_variance = TRUE,
                     estimate_prior_mixture_weights = TRUE,
                     standardize = TRUE, verbose = FALSE)
cat("Fixed weights ELBO:", tail(fit_fixed_w$elbo, 1), "\n")
cat("Estimated weights ELBO:", tail(fit_est_w$elbo, 1), "\n")
cat("(Higher ELBO is better)\n")
```

The weight estimation method can be set via `mixture_weight_method`:
`"mixsqp"` (default, uses the mixsqp package for fast optimization)
or `"EM"`. The former might be slower per iteration but the latter 
might take longer to converge.