---
title: Engineering verification
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Engineering verification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette verifies the internal consistency of the mvSuSiE
implementation through two engineering checks: (1) the multivariate
model reduces to univariate SuSiE when $R = 1$, and (2) the mixture
prior code path agrees with the single-matrix code path when $K = 1$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

```{r simulate-data}
data(simdata)
X <- simdata$raw$X
n <- nrow(X); p <- ncol(X); r <- 10

B <- matrix(0, p, r)
causal <- sort(sample(p, 3))
for (j in causal) B[j, ] <- rnorm(r, 0, 0.5)
Y <- X %*% B + matrix(rnorm(n * r), n, r)
```

## R = 1 reduces to univariate SuSiE

When there is only one outcome ($R = 1$), the multivariate model
should give the same result as univariate SuSiE. This verifies that
the multivariate code path correctly handles the scalar case.

```{r r1-comparison}
Y1 <- Y[, 1]

# Univariate susie
fit_susie <- susie(X, Y1, L = 10,
                   scaled_prior_variance = 0.2,
                   estimate_prior_variance = TRUE,
                   standardize = TRUE,
                   max_iter = 100, tol = 1e-4)

# Multivariate mvsusie with R=1
fit_mv1 <- mvsusie(X, Y1, L = 10,
                   prior_variance = 0.2,
                   estimate_prior_variance = TRUE,
                   standardize = TRUE,
                   max_iter = 100, tol = 1e-4, verbose = FALSE)

cat("susie CS:", length(fit_susie$sets$cs), "\n")
cat("mvsusie (R=1) CS:", length(fit_mv1$sets$cs), "\n")
cat("max |PIP diff|:", max(abs(fit_susie$pip - fit_mv1$pip)), "\n")
```

```{r r1-plot, fig.height=3, fig.width=3.5}
pdat <- data.frame(susie = fit_susie$pip, mvsusie = fit_mv1$pip)
ggplot(pdat, aes(x = susie, y = mvsusie)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  labs(x = "PIP (susie)", y = "PIP (mvsusie, R=1)") +
  theme_cowplot(font_size = 12)
```

The PIPs agree at machine precision, confirming the $R = 1$ reduction.

## Single-effect regression agreement

At the core of both SuSiE and mvSuSiE is the single-effect
regression (SER) step. When $R = 1$, mvSuSiE's SER should
give the same Bayes factors and posterior weights as SuSiE's.

```{r ser-comparison}
# Compare Bayes factors from the first single effect
cat("susie lbf (first effect):", round(fit_susie$lbf[1], 4), "\n")
cat("mvsusie lbf (first effect):", round(fit_mv1$lbf[1], 4), "\n")

# Compare alpha (posterior weights)
alpha_susie <- fit_susie$alpha[1, ]
alpha_mv <- fit_mv1$alpha[1, ]
cat("max |alpha diff| (first effect):",
    max(abs(alpha_susie - alpha_mv)), "\n")
```

## Mixture prior (K > 1) vs single matrix prior (K = 1)

mvSuSiE supports two code paths for the prior: a single $R \times R$
covariance matrix, and a mixture of $K$ matrices (the "mash" prior).
When $K = 1$, the mixture prior should give the same result as
passing a single matrix directly. This verifies that the mixture
machinery does not introduce any discrepancies.

```{r mixture-vs-single}
U <- 0.2 * diag(r)

# Single matrix prior (K=1 code path)
fit_single <- mvsusie(X, Y, L = 10,
                      prior_variance = U,
                      residual_variance = diag(r),
                      estimate_residual_variance = FALSE,
                      estimate_prior_variance = TRUE,
                      standardize = TRUE,
                      max_iter = 100, tol = 1e-4, verbose = FALSE)

# Mixture prior with K=1 (mash code path)
prior_k1 <- create_mixture_prior(
  list(matrices = list(U), weights = 1)
)
fit_mixture <- mvsusie(X, Y, L = 10,
                       prior_variance = prior_k1,
                       residual_variance = diag(r),
                       estimate_residual_variance = FALSE,
                       estimate_prior_variance = TRUE,
                       standardize = TRUE,
                       max_iter = 100, tol = 1e-4, verbose = FALSE)

cat("Single matrix CS:", length(fit_single$sets$cs), "\n")
cat("Mixture (K=1) CS:", length(fit_mixture$sets$cs), "\n")
cat("max |PIP diff|:", max(abs(fit_single$pip - fit_mixture$pip)), "\n")
```

```{r mixture-vs-single-plot, fig.height=3, fig.width=3.5}
pdat <- data.frame(single = fit_single$pip,
                   mixture = fit_mixture$pip)
ggplot(pdat, aes(x = single, y = mixture)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  labs(x = "PIP (single matrix prior)",
       y = "PIP (mixture prior, K=1)") +
  theme_cowplot(font_size = 12)
```