---
title: Univariate-multivariate agreement
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Univariate-multivariate agreement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette verifies that the multivariate mvSuSiE model reduces
correctly to independent univariate SuSiE models under a diagonal
prior, and that the MASH prior reduces to univariate when $R = 1$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

Diagonal prior reduces to independent univariate
--------------------------------------------------

When the prior covariance is diagonal ($U = \text{diag}(\sigma_1^2,
\ldots, \sigma_R^2)$) and the residual covariance is also diagonal,
the multivariate regression decomposes into $R$ independent univariate
regressions. We verify this by fitting mvSuSiE with a diagonal prior
and comparing to $R$ separate SuSiE fits.

```{r simulate-data}
data(N3finemapping, package = "susieR")
X <- N3finemapping$X
n <- nrow(X); p <- ncol(X); r <- 10

# Simulate independent traits
B <- matrix(0, p, r)
causal <- sort(sample(p, 3))
for (j in causal) B[j, ] <- rnorm(r, 0, 0.5)
Y <- X %*% B + matrix(rnorm(n * r), n, r)
```

```{r fit-multivariate-diagonal}
# Diagonal prior: independent across traits
U_diag <- 0.2 * diag(r)
prior_diag <- create_mixture_prior(
  list(matrices = list(U_diag), weights = 1)
)

# Multivariate fit
fit_mv <- mvsusie(X, Y, L = 10,
                  prior_variance = prior_diag,
                  residual_variance = diag(r),
                  estimate_residual_variance = FALSE,
                  estimate_prior_variance = FALSE,
                  standardize = TRUE,
                  max_iter = 100, tol = 1e-4, verbosity = 0)
```

```{r fit-univariate}
# R independent univariate fits
fits_uv <- lapply(seq_len(r), function(i) {
  susie(X, Y[, i], L = 10,
        scaled_prior_variance = 0.2,
        residual_variance = 1,
        estimate_residual_variance = FALSE,
        estimate_prior_variance = FALSE,
        standardize = TRUE,
        max_iter = 100, tol = 1e-4)
})
```

Compare the posterior inclusion probabilities:

```{r compare-diagonal, fig.height=3.5, fig.width=7}
pip_uv <- sapply(fits_uv, function(f) f$pip)  # p x r
pip_mv <- do.call(cbind, lapply(seq_len(r), function(i) {
  1 - apply(1 - fit_mv$alpha * fit_mv$b1[, , i] /
              (fit_mv$b1[, , i] + 1e-20), 2, prod)
}))

# Use overall PIP for a simpler comparison
pip_uv_overall <- 1 - apply(1 - pip_uv, 1, prod)
pip_mv_overall <- fit_mv$pip

pdat <- data.frame(univariate = pip_uv_overall,
                   multivariate = pip_mv_overall)
ggplot(pdat, aes(x = univariate, y = multivariate)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  labs(x = "PIP (R independent susie)",
       y = "PIP (mvsusie with diagonal prior)") +
  theme_cowplot(font_size = 12)
```

```{r diagonal-agreement}
cat("max |PIP difference|:", max(abs(pip_uv_overall - pip_mv_overall)), "\n")
cat("PIP correlation:", cor(pip_uv_overall, pip_mv_overall), "\n")
```

MASH prior reduces to univariate when R = 1
---------------------------------------------

When there is only one outcome ($R = 1$), the multivariate model
should give the same result as univariate SuSiE.

```{r r1-comparison}
Y1 <- Y[, 1]

# Univariate susie
fit_susie <- susie(X, Y1, L = 10,
                   scaled_prior_variance = 0.2,
                   estimate_prior_variance = TRUE,
                   standardize = TRUE,
                   max_iter = 100, tol = 1e-4)

# Multivariate mvsusie with R=1
fit_mv1 <- mvsusie(X, Y1, L = 10,
                   prior_variance = 0.2,
                   estimate_prior_variance = TRUE,
                   standardize = TRUE,
                   max_iter = 100, tol = 1e-4, verbosity = 0)

cat("susie CS:", length(fit_susie$sets$cs), "\n")
cat("mvsusie (R=1) CS:", length(fit_mv1$sets$cs), "\n")
cat("max |PIP diff|:", max(abs(fit_susie$pip - fit_mv1$pip)), "\n")
```

```{r r1-plot, fig.height=3, fig.width=3.5}
pdat <- data.frame(susie = fit_susie$pip, mvsusie = fit_mv1$pip)
ggplot(pdat, aes(x = susie, y = mvsusie)) +
  geom_point(shape = 20, size = 1.5, color = "royalblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  labs(x = "PIP (susie)", y = "PIP (mvsusie, R=1)") +
  theme_cowplot(font_size = 12)
```

Single-effect regression agreement
------------------------------------

At the core of both SuSiE and mvSuSiE is the single-effect
regression (SER) step. When $R = 1$, mvSuSiE's SER should
give the same Bayes factors and posterior weights as SuSiE's.

```{r ser-comparison}
# Compare Bayes factors from the first single effect
cat("susie lbf (first effect):", round(fit_susie$lbf[1], 4), "\n")
cat("mvsusie lbf (first effect):", round(fit_mv1$lbf[1], 4), "\n")

# Compare alpha (posterior weights)
alpha_susie <- fit_susie$alpha[1, ]
alpha_mv <- fit_mv1$alpha[1, ]
cat("max |alpha diff| (first effect):",
    max(abs(alpha_susie - alpha_mv)), "\n")
```

These results confirm that the multivariate implementation correctly
reduces to the univariate case, providing confidence in the more
general multivariate computations.
