---
title: Interpreting and visualizing mvSuSiE results
author: Gao Wang
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting and visualizing mvSuSiE results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates how to interpret the output of a fitted
mvSuSiE model, including credible sets, posterior inclusion
probabilities (PIPs), effect estimates, and local false sign rates
(LFSR). We also show the built-in visualization functions.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r load-pkgs, message=FALSE}
library(mvsusieR)
library(ggplot2)
library(cowplot)
set.seed(1)
```

We use the built-in `simdata` data set, which contains realistic
genotypes and simulated phenotypes for 20 traits with three causal
SNPs.

```{r fit-model}
data(simdata)
X <- simdata$raw$X
Y <- simdata$raw$Y
cat(sprintf("Data: %d samples, %d SNPs, %d traits\n",
            nrow(X), ncol(X), ncol(Y)))

prior <- create_mixture_prior(list(matrices = simdata$par$U,
                                   weights = simdata$par$w),
                              null_weight = 0)
fit <- mvsusie(X, Y, standardize = TRUE,
               prior_variance = prior,
               residual_variance = simdata$par$V,
               estimate_prior_variance = TRUE,
               tol = 0.01)
```

Credible sets
-------------

mvSuSiE identifies causal SNPs through 95% credible sets (CSs).
Each CS is a set of variables that, with 95% posterior probability,
contains the causal variant for one of the single effects.

```{r credible-sets}
cat("Number of credible sets:", length(fit$sets$cs), "\n\n")
fit$sets$cs
```

CS purity measures the minimum absolute correlation between any pair
of SNPs in the CS. High purity (close to 1) means the CS cannot be
further refined because the remaining SNPs are tightly correlated.

```{r cs-purity}
fit$sets$purity
```

Posterior inclusion probabilities
----------------------------------

The PIP for each SNP gives the posterior probability of being causal
(included in at least one single effect).

```{r pip}
cat("PIP range:", round(range(fit$pip), 6), "\n")
cat("SNPs with PIP > 0.5:", sum(fit$pip > 0.5), "\n")
```

PIP and effect plots
--------------------

`mvsusie_plot()` produces two key visualizations: a PIP plot showing
which SNPs are in credible sets, and an effect plot showing the
estimated per-trait effects at the sentinel SNPs.

```{r pip-plot, fig.height=2, fig.width=6}
out <- mvsusie_plot(fit)
out$pip_plot
```

The effect plot shows which traits are affected by each causal SNP.
Point size indicates effect magnitude and color indicates sign
(red = positive, blue = negative). Only effects with local false
sign rate (LFSR) below the cutoff (default 0.01) are shown.

```{r effect-plot, fig.height=3.25, fig.width=2.5}
out_cs <- mvsusie_plot(fit, add_cs = TRUE)
out_cs$effect_plot
```

The first CS identifies a SNP affecting most traits, while the third
CS identifies a SNP affecting only one trait.

Local false sign rate
---------------------

The LFSR for a variable-trait pair gives the posterior probability that
the estimated sign of the effect is wrong. Small LFSR values indicate
confident effect identification. LFSR can be thought of as an
alternative to a *p*-value, but is typically slightly more conservative.

```{r lfsr}
cat("LFSR matrix dimensions:", dim(fit$lfsr), "\n")
cat("(rows = SNPs, columns = traits)\n\n")

# Show LFSR for SNPs in credible sets
for (cs_name in names(fit$sets$cs)) {
  sentinel <- fit$sets$cs[[cs_name]][which.max(
    fit$pip[fit$sets$cs[[cs_name]]])]
  n_sig <- sum(fit$lfsr[sentinel, ] < 0.01)
  cat(sprintf("%s sentinel (SNP %d): %d traits with LFSR < 0.01\n",
              cs_name, sentinel, n_sig))
}
```

Single-effect LFSR gives the LFSR for each single effect (credible
set) across traits:

```{r single-effect-lfsr}
cat("Single-effect LFSR dimensions:", dim(fit$single_effect_lfsr), "\n")
cat("(rows = effects, columns = traits)\n")
```

Outcome-specific PIPs
---------------------

The overall PIP tells us whether a SNP affects *any* trait. The
outcome-specific PIP tells us the probability that a SNP affects
a *specific* trait. This is computed from the mixture weights and
prior structure.

```{r per-outcome-pip}
pip_po <- mvsusie_get_pip_per_outcome(fit)
cat("Per-outcome PIP matrix:", dim(pip_po), "\n")
```

```{r per-outcome-pip-plot, fig.height=3, fig.width=5}
# Compare overall PIP vs outcome-specific PIP for a few traits
traits_to_show <- c(1, 5, 10, 15)
pdat <- do.call(rbind, lapply(traits_to_show, function(t) {
  data.frame(overall_pip = fit$pip,
             outcome_pip = pip_po[, t],
             trait = paste0("Trait ", t))
}))
ggplot(pdat, aes(x = overall_pip, y = outcome_pip)) +
  geom_point(shape = 20, size = 1, color = "royalblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  facet_wrap(~trait) +
  labs(x = "Overall PIP", y = "Outcome-specific PIP") +
  theme_cowplot(font_size = 10)
```

Per-effect coefficient estimates
---------------------------------

The `b1_rescaled` array contains the posterior mean effect estimates
for each single effect on the original scale. This is an $L \times
(J+1) \times R$ array (the first column is the intercept).

```{r effect-estimates, fig.height=2.5, fig.width=3.5}
l <- 1
sentinel <- fit$sets$cs$L1[which.max(fit$pip[fit$sets$cs$L1])]
pdat <- data.frame(
  true = simdata$Btrue[sentinel, ],
  estimated = fit$b1_rescaled[l, sentinel + 1, ],
  lfsr = fit$single_effect_lfsr[l, ]
)
ggplot(pdat, aes(x = true, y = estimated,
                 color = log10(lfsr))) +
  geom_point(shape = 20, size = 2.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  scale_color_gradient2(low = "lightskyblue", mid = "gold",
                        high = "orangered", midpoint = -40) +
  labs(x = "True coefficient", y = "mvSuSiE estimate") +
  theme_cowplot(font_size = 12)
```

The mvSuSiE estimates align well with the true coefficients, and
traits with significant effects (small LFSR) are correctly identified.
